---
title: "aging_mouse_ev_proteome_figures"
author: "Kristine Tsantilas, PhD"
date: "`r Sys.Date()`"
output: 
  html_document:
      code_folding: hide
      toc: true
      toc_float:
        collapsed: false
---

# Overview {.tabset .tabset-fade}

$~$

## Experiments

Aging Cohort

* Sample preparation

  - Mag-Net enrichment of circulating particles (Denoted EVs going forward) from 20 $\mu$L of mouse plasma.
  
  - Denaturation, reduction, alkylation, and protein aggregation and capture (PAC) and tryptic digestion on strong anion exchange (SAX) beads.
  
  - 24 minute LC gradients using Vanquish Neo (Thermo) coupled to a Thermo Orbitrap Astral. All data collected with data-independent acquisition MS (DIA-MS) 
  
  - Resulting data was analyzed using gas-phase fractionated libraries comprised of a pool of the experimental samples collected in separate narrow-window runs.


$~$


Enrichment/Depletion Experiment

* Sample preparation

  - Processed 1 $\mu$L of neat pooled mouse plasma and performed Mag-Net enrichment from 100 $\mu$L of the same mouse plasma.
  
  - In the Mag-Net prep, denaturation, reduction, alkylation, and protein aggregation and capture (PAC), tryptic digestion were done on strong anion exchange (SAX) beads. For the neat plasma, denaturation, reduction, and alkylation were done before precipitation onto hydroxyl beads and subsequent tryptic digestion
  
  - 60 minute LC gradients using EasyLC 1200 coupled to a Thermo Orbitrap Exploris 480. All data were collected with data-independent acquisition MS (DIA-MS) 
  
  - Resulting data was analyzed using gas-phase fractionated libraries comprised of a pool of the experimental samples collected in separate narrow-window runs.


<br/><br/>


## Input

Sample data and metadata are needed to run the R markdown. 

* Aging cohort data:

  + Peptide-level and protein-level data for each replicate with different normalization exported from skyline as .csv files.

    - Protein, normalized: proteins_normalized_wide.tsv
  
    - Protein, raw: proteins_unnormalized_wide.tsv
    
    - Peptide, normalized: precursors_normalized_wide.tsv
    
    - Peptide, raw: precursors_unnormalized_wide.tsv
   
  + Metadata
  
    - metadata_wide.tsv
    
    - Manual author categorization of proteins changed differentially in aging by sex: sex_proteins_table_manual_annotations.csv


* Mag-Net species check

  + Nanosight particle counting data

    - Replicate A: IEQC_ReplicateA_1to25_ExperimentSummary_ParticleCounts.csv
    
    - Replicate B: IEQC_ReplicateB_1to25_ExperimentSummary_ParticleCounts.csv
        
  + LC-MS data and metadata related to the analysis of total plasma and EV-enriched Mag-Net fraction
   
   - Fold-change between plasma and Mag-Net Fraction: fold_change_plasma_evs_protein_level.csv # Mouse total plasma and Mag-Net fraction data
   
   - Mouse total plasma and Mag-Net fraction metadata: plasma_ev_replicates.csv
   
   - Targets for enrichment and depletion plots: ev_and_hap_proteins.csv
   

* Additional data relevant to the aging cohort data generated by Mike Riffle in Jupyter Notebook imported to generate figures and supplementary tables.

  + Sex-differential Spearman results: protein-sex-discordant-spearman-rho.csv

  + Sex-differential OLS results: mouse-sex-specific-age-effect-features-ols.csv
  
  + Protein-level regression analysis: protein-age-regression-final-model-coefficients.csv 


<br/><br/>


## Output

This script produces:

* An HTML format R Markdown file that includes the experimental data, code detailed here, and print-outs of figures.

* Figures (as .svg files)

* Various .csv files with subset of code steps


<br/><br/>


## Software Information

* Skyline Daily (64-bit)

* R version: 4.1.2

* R Studio version: 2023.03.0+386 "Cherry Blossom" Release (3c53477afb13ab959aeb5b34df1f10c237b256c3, 2023-03-09) for Windows
Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36 (KHTML, like Gecko) RStudio/2023.03.0+386 Chrome/108.0.5359.179 Electron/22.0.3 Safari/537.36

* Packages required: 

  + RColorBrewer (version 1.1-3)

  + dplyr (version 1.1.2)
  
  + stringr (version 1.5.0)
  
  + tidyr (version 1.3.0)
  
  + matrixStats (version 0.63.0)
  
  + ggplot2 (version 3.4.2)
  
  + ggrepel (version 0.9.3)

  + ggpubr (version 0.6.0)

  + knitr (version 1.43)
  
  + kableExtra (version 1.3.4)

  + patchwork (version 1.1.2)
  
  + scales (version 1.2.1) 
  
<br/><br/>



```{r setup}
# Setting global options for all code chunks
    knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, warning=FALSE, message=FALSE)

    working_directory <- getwd()
    print(working_directory)
    
    
# Setting option for knitting
    knitr::opts_knit$set(echo = TRUE, root.dir = working_directory)
    
    setwd(working_directory)
    getwd()


```

```{r package_installation, include=FALSE}
# BiocManager and associated BioConductor package installation if needed.
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install(version = "3.19")
# 
#     BiocManager::install(c("UniprotR", "ReactomePA", "clusterProfiler", "AnnotationDbi","org.Mm.eg.db", "DOSE"))
    # BiocManager::install("")
    # BiocManager::install("org.Hs.eg.db")
    # BiocManager::install("ComplexHeatmap")

#Download necessary packages for analysis
    library(car)
    library(RColorBrewer)
    library(dplyr)
    library(stringr)
    library(tidyr)
    library(matrixStats)
    library(ggplot2)
    library(purrr)
    library(ggnewscale)
    library(ggplotify)
    library(ggrepel)
    library(ggpubr)
    library(knitr)
    library(data.table)
    library(pheatmap)
    library(grid)
    library(kableExtra)
    library(patchwork)
    library(cowplot)
    library(dendextend)
    library(viridis)
    library(scales)

# Bioconductor packages
    library(UniprotR)
    library(ReactomePA)
    library(AnnotationDbi)
    library(clusterProfiler)
    library(org.Mm.eg.db)
    library(DOSE)


# Check session info
sessionInfo()


```




```{r setup_directories_labels}

# Set-up current date to include in filename output
    (today <- Sys.Date())
    basename <- today %>%
                format(., "%Y%b%d") %>% # with month as a word
                paste0(.,"_")

# Set-up base filenames by dataset to classify filename output
    filenamebase <- "Mouse_EV_"

# Setting up sub-folders in working directory
    directories <- c('rmd_output', 'rmd_output/plots', 'rmd_output/tables')
    
    for(c in directories){
      if(!dir.exists(c)){
        message(paste('Creating directory: ', c))
        dir.create(c)
      }}

```


<br/><br/>

## File Import

Importing files including sample data and metadata.

```{r file_import}

# Figure 1
      # No files used


# Figure 2 - 4 and Figure S2
      pep_unnorm <- read.delim("input\\precursors_unnormalized_wide.tsv", sep = "\t")
      pep_norm <- read.delim("input\\precursors_normalized_wide.tsv", sep = "\t")
      
      prot_unnorm <- read.delim("input\\proteins_unnormalized_wide.tsv", sep = "\t")
      prot_norm <- read.delim("input\\proteins_normalized_wide.tsv", sep = "\t")
      
      metadata <- read.delim("input\\metadata_wide.tsv",sep="\t")




# Figure S1
    # Particle counting distribution of Mag-Net fraction collected 
    # on Malvern Nanosight NS300 (Software NTA 3.4, Build 3.4.4)
      particle_count_A <- read.csv("input\\IEQC_ReplicateA_1to25_ExperimentSummary_ParticleCounts.csv",
                                   header = TRUE, na.strings="#N/A", sep = ",")

      particle_count_B <- read.csv("input\\IEQC_ReplicateB_1to25_ExperimentSummary_ParticleCounts.csv",
                                   header = TRUE, na.strings="#N/A", sep = ",")

      # Renaming unruly column headers
                  names(particle_count_A) <- gsub(x = names(particle_count_A), pattern = "Concentration..particles...ml..", replacement = "Replicate_")  
                  names(particle_count_A) <- gsub(x = names(particle_count_A), pattern = "Concentration..particles...ml.", replacement = "Replicate_0")  
                  
                  names(particle_count_B) <- gsub(x = names(particle_count_B), pattern = "Concentration..particles...ml..", replacement = "Replicate_")
                  names(particle_count_B) <- gsub(x = names(particle_count_B), pattern = "Concentration..particles...ml.", replacement = "Replicate_0")  
    
                  
    # Mouse total plasma and Mag-Net fraction data
      magnet_dia <- read.csv("input\\fold_change_plasma_evs_protein_level.csv",
                             header = TRUE, na.strings="#N/A", sep = ",")

    # Mouse total plasma and Mag-Net fraction metadata
      magnet_reps <- read.csv("input\\plasma_ev_replicates.csv",
                              header = TRUE, na.strings="#N/A", sep = ",") %>%
                     select(Replicate, SampleGroup)

    # Targets for enrichment and depletion plots
      magnet_prot_targets <- read.csv("input\\ev_and_hap_proteins.csv",
                                      header = TRUE, na.strings="#N/A", sep = ",")

      
    # Sex-differential Spearman results by Mike Riffle in Jupyter notebook
      sex_spearman <- read.csv("input\\python_results\\protein-sex-discordant-spearman-rho.csv",
                                      header = TRUE, na.strings="#N/A", sep = ",")
      
    # Sex-differential OLS results by Mike Riffle in Jupyter notebook
      sex_ols <- read.csv("input\\python_results\\mouse-sex-specific-age-effect-features-ols.csv",
                                      header = TRUE, na.strings="#N/A", sep = ",")
# Table S3
    # Manual categorization of proteins changed differential in aging by sex
      manual_prot_cats <- read.csv("input\\sex_proteins_table_manual_annotations.csv", 
                                   #fileEncoding = "Windows-1252",
                                    header = TRUE, na.strings="#N/A", sep = ",") %>%
                          rename("Functional Categories" = "Functional.categories")
      
    # Protein-level regression analysis results by Mike Riffle in Jupyter notebook
      prot_regr_age <- read.csv("input\\python_results\\protein-age-regression-final-model-coefficients.csv",
                                      header = TRUE, na.strings="#N/A", sep = ",")

         
```

</br></br>

# Script Functions {.tabset .tabset-fade}

Various functions were defined and used in this R Markdown file to perform analyses and generate figures.

$~$

## Tidying

Tidying DIA-MS results from aging cohort that were used to generate multiple figures as wide-format dataframe, long-format dataframe, and data matrix.

```{r function_tidying_input}

generate_long_dia_meta <- function(dataframe, metadata_dataframe){

    # Extracting uniprot first_accession numbers
        dataframe <- dataframe %>%
                    mutate(first_accession = protein,
                           first_accession = str_extract(first_accession,"(?<=\\|)[^|]+(?=\\|)"),
                           prot_id = protein,
                           prot_id = str_extract(prot_id,"(?<=\\|)[^|]+(?=_)"))
    
    # Pivoting to long-format and cleaning-up replicate names to "SampleID_RunORder" by removing common prefix
        dataframe_long <- dataframe %>%
                           pivot_longer(
                           cols = starts_with("X06_01_01_NeoV_PepMap5mm_PepSep8cmFossil_Astral_24min_Churchill_"),
                           names_to = "replicate",
                           names_prefix = "X06_01_01_NeoV_PepMap5mm_PepSep8cmFossil_Astral_24min_Churchill_",
                           values_to = "log2bc_area",
                           values_drop_na = FALSE) %>%
                           mutate(replicate = str_replace_all(replicate, "_\\d+$", ""))
    
    # Merging long-format peptide data with metadata
        dataframe_meta_long <- full_join(dataframe_long, metadata_dataframe, by = "replicate")
        
        dataframe_meta_long = dataframe_meta_long[order(dataframe_meta_long$replicate),]

        return(dataframe_meta_long)

    }


tidy_wide_colnames <- function(dataframe){

    # Extracting uniprot first_accession numbers
        dataframe <- dataframe %>%
                    mutate(first_accession = protein,
                           first_accession = str_extract(first_accession,"(?<=\\|)[^|]+(?=\\|)"),
                           prot_id = protein,
                           prot_id = str_extract(prot_id,"(?<=\\|)[^|]+(?=_)"))
    
    # Standardizing replicate name
        # Remove common prefix
        names(dataframe) <- sub("X06_01_01_NeoV_PepMap5mm_PepSep8cmFossil_Astral_24min_Churchill_", "", names(dataframe))
    
        # Remove run order designation
        new_col_names <- colnames(dataframe) %>%
                                  str_replace_all("_\\d+$", "")
    
        # Assigning new column names
        colnames(dataframe) <- new_col_names
        
        # Reorder by column names to match metadata
        dataframe = dataframe[, order(colnames(dataframe))]
        
        return(dataframe)
    
    }


## DATA MATRIX ##
    generate_pep_matrix_dia <- function(dataframe){
    
    # Generate row names    
        matrix_row_names = paste0(dataframe$modifiedSequence, "@", dataframe$protein)
        
    # Remove columns that are not DIA data
        dataframe = dataframe %>%
                    select(-c(protein, modifiedSequence,precursorCharge, prot_id, first_accession))
    
    # Reorder by column names to match metadata
        dataframe = dataframe[, order(colnames(dataframe))]

    #Convert to numeric matrix and add the rownames set earlier
        data_matrix = as.matrix(dataframe)
        data_matrix = apply(data_matrix,2,as.numeric)
        rownames(data_matrix) = matrix_row_names
        
        return(data_matrix)

    }

    generate_prot_matrix_dia <- function(dataframe){
    
    # Generate row names    
        matrix_row_names = paste0(dataframe$protein)
        
    # Remove columns that are not DIA data
        dataframe = dataframe %>%
                    select(-c(protein, prot_id, first_accession))
    
    # Reorder by column names to match metadata
        dataframe = dataframe[, order(colnames(dataframe))]

    #Convert to numeric matrix and add the rownames set earlier
        data_matrix = as.matrix(dataframe)
        data_matrix = apply(data_matrix,2,as.numeric)
        rownames(data_matrix) = matrix_row_names
        
        return(data_matrix)

    }    

```

<br/><br/>

Defined functions with the input parameter **x** to remove Log~2~ normalization where needed.

```{r function_denorm_log}
remove_log2 <- function(x) 2^(x)

```


<br/><br/>


```{r function_get_uniprot_acc}

# Extracting uniprot first_accession numbers and protein ID. Uniprot is used to get Entrez ID for reactome pathway analysis in later steps.
get_1st_uniprot <- function(dataframe){
    dataframe_with_ids <- dataframe %>%
                          mutate(first_accession = protein,
                                 first_accession = str_extract(first_accession,"(?<=\\|)[^|]+(?=\\|)"),
                                 prot_id = protein,
                                 prot_id = str_extract(prot_id,"(?<=\\|)[^|]+(?=_)"))
    return(dataframe_with_ids)
    }

```

<br/><br/>

## Figure 1 Timeline

Defined functions to put together the colored gradients in the timeline of Figure 1A and to perform the Reactome enrichment analysis and the associated plot in Figure 1B.

* fig1_gradients

  - dataframe: Dataframe containing sequential data for timeline gradient
  
  - y_parameter: Column name (months) to set axes numbers
  
  - fill_parameter: Column name (months) to set fill by numbers
  
  - median_lifespan: Median lifespan (Differs by sex)
  
  - color_low: Darker gradient color correlating with younger (lower) ages
  
  - color_mid: Medium gradient color correlating with middle ages
  
  - color_high: Lighter gradient color correlating with older (higher) ages
        
```{r function_fig1_gradient}

# Setting-up a function to make male and female gradients
fig1_gradients <- function(dataframe, y_parameter, fill_parameter, median_lifespan, color_low, color_mid, color_high){
  
baseplot_gradient <- (ggplot(dataframe) +
                      geom_tile(aes(x = 0.5, y = y_parameter, fill = fill_parameter)) +
                      scale_x_continuous(limits=c(0,1),breaks=1)+
                      scale_fill_gradient2(low = color_low, mid = color_mid, high = color_high, midpoint = 17.5) +
                      coord_flip() +
                      theme_void() +
                      theme(plot.margin = margin(0, 0.2, 0, 0.2, "in"),  # t, r, b, l
                            panel.spacing = unit(0, "pt"),
                            legend.position = "none",#,
                            axis.title = element_blank(),
                            axis.text = element_blank(),
                            axis.ticks = element_blank(),
                            axis.ticks.length = unit(0, "pt"),
                            plot.background = element_blank()) +
                      annotate("pointrange", x = 0.5, xmin = 0, xmax = 1, y = median_lifespan, shape=18, alpha = 1, color = "black", linewidth = 0.25, linewidth = 0.25))

}

```


<br/><br/>

## CV Density - Samples

Defined function with the following input parameters to generate density plots of CVs from two groups


```{r function_cv_dens_sample}

plot_cv_dens <- function(dataframe, analyte, grouping_vars, title_string, colorvector, labels, legtitle, xaxmax, yaxmax){
    cv_density <- (ggplot((dataframe %>%
                      select(c(analyte, SampleGroup, norm_cond, log2bc_area)) %>%
                      group_by(across(all_of(grouping_vars))) %>%
                      summarise(mean_area = mean(remove_log2(log2bc_area)),
                                sd_area = sd(remove_log2(log2bc_area)),
                                cv = ((sd_area / mean_area)*100),
                                .groups = 'drop')), 
              aes(fill = SampleGroup,
                  x = cv)) + 
              geom_density(adjust=1.5, alpha=.5) +
                              theme_minimal()+
                              theme(plot.background = element_blank(),
                                    panel.grid.major.x = element_blank(),
                                    panel.grid.major.y = element_line(color="azure3"),
                                    panel.grid.minor.x = element_blank(),
                                    panel.grid.minor.y = element_blank(),
                                    axis.text = element_text(size=15, color = "black"),
                                    axis.title.y = element_text(size = 16, color = "black", 
                                                                margin = margin(t = 0, r = 0, b = 0, l = 0)),
                                    axis.title=element_text(size=17, color = "black"),
                                    plot.title=element_text(size=17, color = "black"),
                                    legend.position = "right",
                                    legend.title = element_text(size=17, color = "black"),
                                    legend.text = element_text(size=15, color = "black"),
                                    ) +
                              geom_hline(yintercept=0, linetype="solid", color = "black", linewidth=0.5) +
                              geom_vline(xintercept=0, linetype="solid", color = "black", linewidth=0.5) +
                              ggtitle(title_string)+
                              scale_fill_manual(values = colorvector, labels = labels) +
                              scale_color_manual(values = colorvector) +
                              scale_x_continuous(name="Percent CV", limits = c(0,xaxmax)) +
                              scale_y_continuous(name="Density", limits = c(0,(yaxmax+(yaxmax*0.1)))) +
                              guides(fill=guide_legend(title=legtitle)) )
    return(cv_density)
    }


```


<br/><br/>


## CV Density - QCs

Defined function with the following input parameters to generate density plots of CVs from two groups


```{r function_cv_dens_qc}

plot_cv_dens_qc <- function(dataframe, analyte, grouping_vars, title_string, colorvector, labels, legtitle, xaxmax, yaxmax){

    cv_density <- (ggplot((dataframe %>%
                      filter(protein != "sp|P00924|ENO1_YEAST") %>%
                      select(c(analyte, SampleGroup, log2bc_area, norm_cond)) %>%
                      group_by(across(all_of(grouping_vars))) %>%
                      summarise(mean_area = mean(remove_log2(log2bc_area)),
                                sd_area = sd(remove_log2(log2bc_area)),
                                cv = ((sd_area / mean_area)*100),
                                .groups = 'drop')), 
              aes(fill = norm_cond,
                  x = cv)) + 
              geom_density(adjust=1.5, alpha=.5) +
                              theme_minimal()+
                              theme(plot.background = element_blank(),
                                    panel.grid.major.x = element_blank(),
                                    panel.grid.major.y = element_line(color="azure3"),
                                    panel.grid.minor.x = element_blank(),
                                    panel.grid.minor.y = element_blank(),
                                    axis.text = element_text(size=15, color = "black"),
                                    axis.title.y = element_text(size = 16, color = "black", 
                                                                margin = margin(t = 0, r = 0, b = 0, l = 0)),

                                    axis.title=element_text(size=17, color = "black"),
                                    plot.title=element_text(size=17, color = "black"),
                                    legend.position = "right",
                                    legend.title = element_text(size=17, color = "black"),
                                    legend.text = element_text(size=15, color = "black"),
                                    ) +
                              geom_hline(yintercept=0, linetype="solid", color = "black", linewidth=0.5) +
                              geom_vline(xintercept=0, linetype="solid", color = "black", linewidth=0.5) +
                              ggtitle(title_string)+
                              scale_fill_manual(values = colorvector) +
                              scale_color_manual(values = colorvector) +
                              scale_x_continuous(name="CV", limits = c(0,xaxmax)) +
                              scale_y_continuous(name="Density", limits = c(0,(yaxmax+(yaxmax*0.1)))) +
                              guides(fill=guide_legend(title=legtitle)) )
    return(cv_density)
    }


```



<br/><br/>

## PCA - Continuous

Defined functions with the following input parameters to put together PCA plots of continuous variables:

* plot_pca_continuous

    - data_matrix_for_pca: Data matrix that will be used to generate the principal component analysis generated with prcomp. Resulting table include the standard deviation, proportion of variance, and cumulative proportion of variance of all of the principal components.
  
  - metadata_type_label: Text that will be the title of the figure legend.
  
  - metadata_column_color: Variable distinguishing groups in comparison. Should be "dataframe$metadata_plotted".
  
  - metadata_column_color: Variable distinguishing groups in comparison.
  
  - color_palette: Pre-defined color palette vector where "Variable" = "Color" for the selected group is set using scale_colour_gradientn().
  
  - legend_title_lower_margin: Alter spacing between legend title and legend key for better visuals.
  
```{r function_pca_continuous}

plot_pca_continuous <- function(data_matrix_for_pca, metadata_type_label, metadata_column_color, color_palette, legend_title_lower_margin){       
                      print(paste0(metadata_type_label, " PCA: Step 1/4: PCA analysis"))
                      # Run Principal Component Analysis
                          pca_result <- prcomp(t(data_matrix_for_pca), scale. = TRUE)
                    
                      print(paste0(metadata_type_label, " PCA: Step 2/4: Generate dataframe"))
                      # Create a dataframe with PCA results
                          pca_df <- as.data.frame(pca_result$x)
                          pca_result_df <- summary(pca_result)
                      print(paste0(metadata_type_label, " PCA: Step 3/4: Extract PC1 and PC2 variance"))
                      
                          variance_PC1 <- (pca_result_df$importance[2, "PC1"] * 100) # Multiply by 100 to get percentage
                          variance_PC2 <- (pca_result_df$importance[2, "PC2"] * 100)

                      print(paste0(metadata_type_label, " PCA: Step 4/4: Produce and return plot"))
                      
                      plotting_pca <- (ggplot(pca_df, aes(x = PC1, y = PC2, fill = metadata_column_color)) +
                              geom_point(shape = 21, size = 3, color = "#000000") +
                              # stat_ellipse() +
                              labs(x = paste0("PC1: ", variance_PC1, "%"),
                                   y = paste0("PC2: ", variance_PC2, "%")) +
                              theme_minimal() +
                              theme(plot.margin = margin(0.21, 0.1, 0.15, 0.1, "in"),  # t, r, b, l
                                    panel.background = element_rect(fill = NA),
                                    panel.border = element_rect(colour = "black", fill=NA, size=0.5),
                                    plot.title = element_blank(),
                                    axis.text.x = element_text(size=16, color = "black"),
                                    axis.text.y = element_text(size=16, color = "black", vjust = 0.5, hjust=1),
                                    axis.title.x = element_text(size = 18, color = "black", margin = margin(t = 0.1, r = 0, b = 0, l = 0, "in")),
                                    axis.title.y = element_text(size = 18, color = "black", margin = margin(t = 0, r = 0.1, b = 0, l = 0, "in")),
                                    legend.title = element_text(size = 18, hjust = 0.5, 
                                                                margin = margin(t = 0, r = 0, b = legend_title_lower_margin, l = 0, "in")),
                                    legend.text = element_text(size = 15),
                                    legend.margin = margin(t = 0, r = 0.1, b = 0, l = 0, "in"),
                                    legend.key.size = unit(0.3, "in")) +
                              scale_fill_gradientn(name = metadata_type_label, colors = color_palette))

plotting_pca
}



```


<br/><br/>

## PCA - Discrete

Defined functions with the following input parameters to put together PCA plots of continuous and discrete variables:

* plot_pca_discrete

  - data_matrix_for_pca: Data matrix that will be used to generate the principal component analysis generated with prcomp. Resulting table include the standard deviation, proportion of variance, and cumulative proportion of variance of all of the principal components.
  
  - metadata_type_label: Text that will be the title of the figure legend.
  
  - metadata_column_color: Variable distinguishing groups in comparison. Should be "dataframe$metadata_plotted".
  
  - color_palette: Pre-defined color palette vector where "Variable" = "Color" for the selected group is set using scale_color_manual().
  
  - legend_title_lower_margin: Alter spacing between legend title and legend key for better visuals.

```{r function_pca_discrete}

plot_pca_discrete <- function(data_matrix_for_pca, metadata_type_label, metadata_column_color, color_palette, legend_title_lower_margin){       
                      print(paste0(metadata_type_label, " PCA: Step 1/4: PCA analysis"))
                      # Run Principal Component Analysis
                          pca_result <- prcomp(t(data_matrix_for_pca), scale. = TRUE)

                    
                      print(paste0(metadata_type_label, " PCA: Step 2/4: Generate PCA dataframe"))
                      # Create a dataframe with PCA results
                          pca_df <- as.data.frame(pca_result$x)
                          pca_result_df <- summary(pca_result)
                      print(paste0(metadata_type_label, " PCA: Step 3/4: Extract PC1 and PC2 variance"))
                      
                          variance_PC1 <- (pca_result_df$importance[2, "PC1"] * 100) # Multiply by 100 to get percentage
                          variance_PC2 <- (pca_result_df$importance[2, "PC2"] * 100)

                      print(paste0(metadata_type_label, " PCA: Step 4/4: Produce and return plot"))
                      
                      plotting_pca <- (ggplot(pca_df, aes(x = PC1, y = PC2, fill = metadata_column_color)) +
                              geom_point(shape = 21, size = 3, color = "#000000") +
                              # stat_ellipse() +
                              labs(x = paste0("PC1: ", variance_PC1, "%"),
                                   y = paste0("PC2: ", variance_PC2, "%")) +
                              theme_minimal() +
                              theme(plot.margin = margin(0.1, 0.1, 0.15, 0.1, "in"),  # t, r, b, l
                                    panel.background = element_rect(fill = NA),
                                    panel.border = element_rect(colour = "black", fill=NA, size=0.5),
                                    plot.title = element_blank(),
                                    axis.text.x = element_text(size=16, color = "black"),
                                    axis.text.y = element_text(size=16, color = "black", vjust = 0.5, hjust=1),
                                    axis.title.x = element_text(size = 18, color = "black", margin = margin(t = 0.1, r = 0, b = 0, l = 0, "in")),
                                    axis.title.y = element_text(size = 18, color = "black", margin = margin(t = 0, r = 0.1, b = 0, l = 0, "in")),
                                    legend.title = element_text(size = 18, hjust = 0.5, 
                                                                margin = margin(t = 0, r = 0, b = legend_title_lower_margin, l = 0, "in")),
                                    legend.text = element_text(size = 15),
                                    legend.margin = margin(t = 0, r = 0.1, b = 0, l = 0, "in"),
                                    legend.key.size = unit(0.3, "in")) +
                              scale_fill_manual(name = metadata_type_label, values = color_palette))

plotting_pca
}




```


<br/><br/>

## Reactome Pathway Analysis

* gene_id_reactome_analysis

  - dataframe_and_Column: Dataframe and associated column containing Uniprot first_accession numbers.
  
* reactome_dotplot_count

  - reactome_data: Output dataframe from the previous function "gene_id_reactome_analysis" which produces Reactome enrichment analysis
  
  - categories_to_display: Numeric value of the number of pathways to display. Ordered by Gene Ratio.
  
  - midpoint_value: Numeric value - suggest setting near midpoint of the displayed p-values to optimize the p-value color gradient.
  
  - panel_tag: Character setting the plot tag for the final figure

```{r functions_reactome_analysis_dotplot}

#################################################################################
#################################################################################
## 
##       Function to perform reactome pathway enrichment using the 
##       ReactomePA package. Start with a dataframe containing a 
##       column of protein first_accession numbers. 

    gene_id_reactome_analysis <- function(dataframe_and_Column, data_summary){

    # Extract the column with UniProt first_accession Numbers
        uniprot_ids <- dataframe_and_Column

            # Convert protein first_accession number to Gene Symbol and EntrezID (The latter will be used in ReactomePA)
        gene_ids <- bitr(uniprot_ids, fromType="UNIPROT", toType=c("SYMBOL", "ENTREZID"), OrgDb="org.Mm.eg.db")
    # Set variable of Entrez gene IDs for Reactome analysis
        genes <- gene_ids$ENTREZID
    
    # Running Reactome Pathway Enrichment    
        reactome_result <- enrichPathway(gene = genes, organism = "mouse", pAdjustMethod = "BH", pvalueCutoff = 0.05, readable = TRUE)

    # Calculating the gene ratio, backgroun ratio, and odds ratio
        reactome_result_oddsratio <- reactome_result %>%
                                     mutate(GeneRatioNum = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1]) / as.numeric(x[2])),
                                     BgRatioNum = sapply(strsplit(BgRatio, "/"), function(x) as.numeric(x[1]) / as.numeric(x[2])))
        # Calculate the Odds Ratio
        reactome_result_oddsratio <- reactome_result_oddsratio %>%
                                     mutate(OddsRatio = ((GeneRatioNum / (1 - GeneRatioNum)) / (BgRatioNum / (1 - BgRatioNum))))
        
        reactome_result_oddsratio <- reactome_result_oddsratio %>%
                                     arrange(p.adjust)

        
    # Exporting the reactome result generated by enrichPathway() as a .csv file
        write.csv(as.data.frame(reactome_result_oddsratio), file = paste0("rmd_output\\tables\\",basename,filenamebase,"reactome_pathway_enrichment_results",data_summary,".csv"))
    
    # Returning the reactome data as output to use for plotting later
        return(reactome_result_oddsratio)        

    }
        
        
#################################################################################
#################################################################################
## 
##       Dot plot of Reactome results generated in the above user-generated
##       R function "gene_id_reactome_analysis". Set how many categories to
##       display and a midpoint number for the displayed p-values to optimize
##       the p-value color gradient.
##       

reactome_dotplot_count <- function(reactome_data, categories_to_display, midpoint_value, panel_tag, color_high, color_mid, color_low, text_limit, yaxis_spacing){
    # Generate inital barplot to plug into ggplot interface
        dotplot_data <- dotplot(reactome_data, showCategory = categories_to_display)
        
        y_ax_title <- paste0(expression("-Log"[10]~"p-value"))
    # Set-up ggplot object    
        reactome_dot_count <- (ggplot(dotplot_data$data, aes(x = GeneRatio, y = Description)) +
                              geom_point(aes(size = Count, color = -log(p.adjust, base = 10))) +
                              theme_minimal() +
                              theme(
                                axis.text.x = element_text(angle = 45, hjust = 1),
                                axis.text.y = element_text(size = 11, hjust = 1),
                                axis.title.x = element_text(size = 14, face = "bold"),
                                axis.title.y = element_text(size = 14, face = "bold")
                              ) +
                              xlab("Gene Ratio") +
                              ylab("Reactome Pathways") +
                              labs(tag = panel_tag) +
                              scale_y_discrete(expand = c(0.1, yaxis_spacing), labels = function(y) str_wrap(y, width = text_limit)) +
                              scale_color_gradient2(name = y_ax_title, midpoint = midpoint_value, 
                                                    low = color_low, mid = color_mid, high = color_high))

        return(reactome_dot_count)        

}

```


<br/><br/>

## Spearman Correlation

Defined functions with the following input parameters to produce dataframes with spearman correlation of sample data:

* corrspear_function_ageVSabundance

  - dataframe: 

  - analyte_type: 

* corrspear_filtering_pval:

  - dataframe: 

  - analyte_type:  
  
  - adjusted_p_value_cutoff: 

* corrspear_filtering_pval_rho:

  - dataframe
  
  - analyte_type
  
  - adjusted_p_value_cutoff
  
  - lower_spearman_rho
  
  - upper_spearman_rho

```{r functions_spearman_correlation}


corrspear_function_ageVSabundance <- function(dataframe, analyte_type){
        print(paste0(analyte_type, ": Spearman Correlation Function - 1/4: Remove IEQC and IBQC controls and yeast enolase controls from analysis"))
        # Generate a filtered dataframe that removes the controls and yeast enolase
        subdata <- (dataframe %>%
                   filter(SampleGroup != "IEQC" & SampleGroup != "IBQC") %>%
                   filter(protein != "sp|P00924|ENO1_YEAST"))

        
        print(paste0(analyte_type, ": Spearman Correlation Function - 2/4: Spearman Correlation using cor.test() and BH correction for multiple hypothesis testing using p.adjust(method = BH)"))
        # Perform the grouped Spearman correlation and extract results into a dataframe
        output_df <- (subdata %>%
                      reframe(statistic = cor.test(as.numeric(Age_months), as.numeric(log2bc_area), method = "spearman", alternative = "two.sided",
                                           exact = FALSE, conf.level = 0.95, continuity = FALSE)$statistic,
                      estimate = cor.test(as.numeric(Age_months), as.numeric(log2bc_area), method = "spearman", alternative = "two.sided",
                                          exact = FALSE, conf.level = 0.95, continuity = FALSE)$estimate,
                      p_value = cor.test(as.numeric(Age_months), as.numeric(log2bc_area), method = "spearman", alternative = "two.sided",
                                         exact = FALSE, conf.level = 0.95, continuity = FALSE)$p.value) %>%
                  
                      # Calculate R-squared and adjust p-values for multiple testing using BH correction
                                mutate(rSquared = estimate^2,
                                       p_val_adj = p.adjust(p_value, method = "BH")))

        print(paste0(analyte_type, ": Spearman Correlation Function - 3/4: Exporting output as tsv"))
        write.csv(output_df, paste("rmd_output\\tables\\",basename,filenamebase,"corr_spear_",analyte_type,"_all_pAdjust.csv", sep=""), 
                  quote=FALSE, row.names=FALSE)

               
        print(paste0(analyte_type, ": Spearman Correlation Function - 4/4: End of correlation extraction"))
        return(output_df)
}


  corrspear_filtering_pval <- function(dataframe, analyte_type, adjusted_p_value_cutoff){       
    print(paste0(analyte_type, ": Filtering Function - 1/3: p-value filtering to ",adjusted_p_value_cutoff," (BH-adjusted p-value)"))
    #Generate file with p-value filtering after BH correction
        pval_filtered_subdata <- (dataframe %>%
                                 filter(., p_val_adj < adjusted_p_value_cutoff))

        
        print(paste0(analyte_type, ": Filtering Function - 2/3: Exporting output as tsv"))

        write.csv(pval_filtered_subdata, paste("rmd_output\\tables\\",basename,filenamebase,"corr_spear_",analyte_type,"_pAdjust.csv", 
                                               sep=""), quote=FALSE, row.names=FALSE)

        
        print(paste0(analyte_type, ": Filtering Function - 3/3: End of filtering"))

        return(pval_filtered_subdata)
    }
 

  corrspear_filtering_pval_rho <- function(dataframe, analyte_type, adjusted_p_value_cutoff, lower_spearman_rho, upper_spearman_rho){       
        print(paste0(analyte_type, ": Filtering Function - 1/2: p-value filtering coupled with Rho cutoffs ",lower_spearman_rho,"and",upper_spearman_rho))
        pval_rho_filtered_subdata <- (dataframe %>%
                                     filter(estimate <= lower_spearman_rho | estimate >= upper_spearman_rho) %>%
                                     filter(., p_val_adj < adjusted_p_value_cutoff))
      
        
        print(paste0(analyte_type, ": Filtering Function - 2/3: Exporting output as tsv"))
        write.csv(pval_rho_filtered_subdata, paste("rmd_output\\tables\\",basename,filenamebase,"corr_spear_",analyte_type,"_pAdjust_rhocutoff.csv", 
                                                 sep=""), quote=FALSE, row.names=FALSE)

        
        print(paste0(analyte_type, ": Filtering Function - 3/3: End of filtering"))

        return(pval_rho_filtered_subdata)
  }         
  
  
```

<br/><br/>

## Protein Correlation Matrix

```{r function_generate_prot_correlation_matrix}


generate_prot_correlation_matrix <- function(dataframe){
    
    # Generate row names    
        matrix_row_names = paste0(dataframe$protein)
        
    # Remove columns that are not DIA data
        dataframe = dataframe %>%
                    select(-c(protein))
    
    # Reorder by column names to match metadata
        dataframe = dataframe[, order(colnames(dataframe))]

    #Convert to numeric matrix and add the rownames set earlier
        data_matrix = as.matrix(dataframe)
        data_matrix = apply(data_matrix,2,as.numeric)
        rownames(data_matrix) = matrix_row_names
        
        return(data_matrix)

    }    


```


<br/><br/>

## Top 5 correlated proteins

```{r function_extract_top5}

extract_top5 <- function(dataframe_indiv, dataframe_corr, ordered_col, trend_id){

 corr_top5_df <- dataframe_corr %>%
                 arrange(ordered_col) %>%
                 slice_head(n = 5)
 
 corr_top5 <- (unique(corr_top5_df$prot_id))
 
 
 corr_top5_labeled <- dataframe_indiv %>%
                      filter(SampleGroup != "IEQC" & SampleGroup != "IBQC") %>%
                      filter(protein != "sp|P00924|ENO1_YEAST") %>%
                      subset(., prot_id %in% corr_top5) 
 corr_top5_labeled$Trend <- trend_id
 
 return(corr_top5_labeled)   
} 


```

<br/><br/>

## Individual Protein Correlation Plots

```{r function_indiv_corr_plots}
generate_indiv_corr_plots <- function(individual_data, correlation_data, targets_corr, list_corr_prot, manually_set_panel_tag, panel_tags) {
    dataset_size <- n_distinct(individual_data$replicate)
    print(paste0("Currently generating plots of linear model of ", dataset_size, " data points"))
    
  for (k in seq_along(targets_corr)) {
    print(targets_corr[k])
    
    # Set a shared color palette
    colpalette_corr <- c("Aging and Senescence" = "darkgray", 
                               "Increases" = "blue", 
                               "Decreases" = "red")

    # Subset the input data for an individual protein
    indiv_subdata <- individual_data[individual_data$protein == targets_corr[k], ]
    corr_subdata <- correlation_data[correlation_data$protein == targets_corr[k], ]
    
    # Skip if no data available and verbalize
    if (nrow(indiv_subdata) == 0 || nrow(corr_subdata) == 0) {
      message(paste("Skipping protein", targets_corr[k], "due to missing data"))
      next
    }

    # Fit a linear regression model to the individual datapoints
    model_linear <- lm(log2bc_area ~ Age_months, data = indiv_subdata)

    # Generate predictions with confidence intervals
    model_linear_ci <- data.frame(Age_months = seq(min(indiv_subdata$Age_months, na.rm = TRUE),
                                                   max(indiv_subdata$Age_months, na.rm = TRUE),
                                                   length.out = dataset_size))
    predictions <- predict(model_linear, model_linear_ci, interval = "confidence", level = 0.95)
    model_linear_ci <- cbind(model_linear_ci, predictions)

    # R-squared for the model
    r_squared <- summary(model_linear)$r.squared

    # Set y-axis limits
    prot_max <- max(indiv_subdata$log2bc_area)
    prot_min <- min(indiv_subdata$log2bc_area)
    ymax <- prot_max + 0.35
    ymin <- prot_min - 0.65
    y_breaks <- seq(ymin, ymax, length.out = 4)
    
    # Create the correlation plot
        print(paste0("Currently plotting ", unique(indiv_subdata$prot_id)))
    
    corr_indiv <- ggplot() +
      geom_line(data = model_linear_ci, aes(x = Age_months, y = fit), 
                color = "#000000", size = 1) +
      geom_ribbon(data = model_linear_ci, aes(x = Age_months, ymin = lwr, ymax = upr), 
                  fill = "#000000", alpha = 0.15) +
      geom_point(data = indiv_subdata, aes(x = Age_months, y = log2bc_area, fill = Trend), 
                 alpha = 0.8, size = 2, shape = 21, color = "#000000") +
      theme_minimal() +
      theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0),
            plot.background = element_blank(),
            panel.grid = element_blank(),
            axis.text = element_text(size = 13.5, color = "black"),
            axis.title = element_text(size = 13.5, color = "black"),
            plot.title = element_text(size = 15, color = "black", hjust = 0.5, margin = margin(t=4, r=0, b=1, l=0)),
            plot.subtitle = element_text(size = 11, face = "italic", color = "black"),
            legend.position = "none") +
      geom_hline(yintercept = ymin, linetype = "solid", color = "black", linewidth = 0.5) +
      geom_vline(xintercept = 4.5, linetype = "solid", color = "black", linewidth = 0.5) +
      scale_fill_manual(values = colpalette_corr) +
      scale_x_continuous(limits = c(4.5, 32)) +
      scale_y_continuous(limits = c(ymin, ymax),
                         breaks = y_breaks,
                         labels = scales::number_format(accuracy = 1)) +
      labs(title = unique(indiv_subdata$prot_id),
           subtitle = bquote(rho == .(round(corr_subdata$estimate, 2)) * "," ~ 
                             p[adj] == .(formatC(corr_subdata$p_val_adj, format = "e", digits = 0)) ))
    #* "," ~ 
    #                         R[lm]^2 == .(round(r_squared, 2))))

    if (manually_set_panel_tag == TRUE) {
    # Only runs when manually_set_panel_tag function parameter is set to TRUE
           corr_indiv <- (corr_indiv + labs(tag = panel_tags[k]) + theme(plot.tag = element_text(size = 25)))
            }
          
    # Store plot output in list      
    list_corr_prot[[k]] <- corr_indiv
  }

  # Return a list of the plots as the output
  return(list_corr_prot)
}

```

<br/><br/>


## Paired Protein Correlation Plots (by sex)

```{r function_paired_sex_corr_plots}
generate_paired_sex_corr_plots <- function(individual_data_f, correlation_data_f, individual_data_m, correlation_data_m, targets_corr, list_corr_prot, manually_set_panel_tag, panel_tags) {
  for (k in seq_along(targets_corr)) {
    dataset_size_f <- n_distinct(individual_data_f$replicate)
    dataset_size_m <- n_distinct(individual_data_m$replicate)
    print(paste0("Currently generating plots of linear models of ", dataset_size_f, " female and ", dataset_size_m, " male data points"))
    
        print(targets_corr[k])
    
    # Set a shared color palette
    colpalette_sex <- c("Female" = "#5D3A9B", "Male"= "#E66100", "Mixed"= "azure4")

    # Subset the input data for an individual protein
    indiv_subdata_f <- individual_data_f[individual_data_f$protein == targets_corr[k], ]
    corr_subdata_f <- correlation_data_f[correlation_data_f$protein == targets_corr[k], ]

    indiv_subdata_m <- individual_data_m[individual_data_m$protein == targets_corr[k], ]
    corr_subdata_m <- correlation_data_m[correlation_data_m$protein == targets_corr[k], ] 


    # Females
        # Fit a linear regression model to the individual datapoints
        model_linear_f <- lm(log2bc_area ~ Age_months, data = indiv_subdata_f)
    
        # Generate predictions with confidence intervals
        model_linear_ci_f <- data.frame(Age_months = seq(min(indiv_subdata_f$Age_months, na.rm = TRUE),
                                                       max(indiv_subdata_f$Age_months, na.rm = TRUE),
                                                       length.out = dataset_size_f))
        predictions_f <- predict(model_linear_f, model_linear_ci_f, interval = "confidence", level = 0.95)
        model_linear_ci_f <- cbind(model_linear_ci_f, predictions_f)
    
        # R-squared for the model
        r_squared_f <- summary(model_linear_f)$r.squared

    # Males
        # Fit a linear regression model to the individual datapoints
        model_linear_m <- lm(log2bc_area ~ Age_months, data = indiv_subdata_m)

    
        # Generate predictions with confidence intervals
        model_linear_ci_m <- data.frame(Age_months = seq(min(indiv_subdata_m$Age_months, na.rm = TRUE),
                                                       max(indiv_subdata_m$Age_months, na.rm = TRUE),
                                                       length.out = dataset_size_m))
        predictions_m <- predict(model_linear_m, model_linear_ci_m, interval = "confidence", level = 0.95)
        model_linear_ci_m <- cbind(model_linear_ci_m, predictions_m)
    
        # R-squared for the model
        r_squared_m <- summary(model_linear_m)$r.squared


    corr_subdata_f$Sex <- "Female"
    corr_subdata_m$Sex <- "Male"
     
    model_linear_f$Sex <- "Female"
    model_linear_m$Sex <- "Male"
    
    model_linear_ci_f$Sex <- "Female"
    model_linear_ci_m$Sex <- "Male"
    
    indiv_subdata <- rbind(indiv_subdata_f, indiv_subdata_m)
    corr_subdata <- rbind(corr_subdata_f, corr_subdata_m)    
    model_linear <- rbind(model_linear_f, model_linear_m) 
    model_linear_ci <- rbind(model_linear_ci_f, model_linear_ci_m) 
   
    print(paste0("Currently plotting ", unique(indiv_subdata$prot_id)))      

    # Skip if no data available and verbalize
    if (nrow(indiv_subdata) == 0 || nrow(corr_subdata) == 0) {
      message(paste("Skipping protein", targets_corr[k], "due to missing data"))
      next
    }

    print(paste0("Currently plotting ", unique(indiv_subdata$prot_id)))
        
    # Set y-axis limits
    prot_max <- max(indiv_subdata$log2bc_area)
    prot_min <- min(indiv_subdata$log2bc_area)
    ymax <- prot_max + 0.35
    ymin <- prot_min - 0.35
    y_breaks <- seq(ymin, ymax, length.out = 4)

    # Create the correlation plot
    corr_indiv <- ggplot() +
      geom_line(data = model_linear_ci, aes(x = Age_months, y = fit, 
                color = Sex), size = 1.25) +
      geom_ribbon(data = model_linear_ci, aes(x = Age_months, ymin = lwr, ymax = upr, 
                fill = Sex), alpha = 0.15) +
      geom_point(data = (indiv_subdata %>%
                         group_by("Sex")), aes(x = Age_months, y = log2bc_area, fill = Sex), 
                 alpha = 0.8, size = 2, shape = 21, color = "#000000") +
      theme_minimal() +
      theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0),
            plot.background = element_blank(),
            panel.grid = element_blank(),
            axis.text = element_text(size = 13.5, color = "black"),
            axis.title = element_text(size = 14, color = "black"),
            plot.title = element_text(size = 15, color = "black", hjust = 0.5),
            plot.subtitle = element_text(size = 11, face = "italic", color = "black"),
            legend.position = "none") +
      geom_hline(yintercept = ymin, linetype = "solid", color = "black", linewidth = 0.5) +
      geom_vline(xintercept = 4.5, linetype = "solid", color = "black", linewidth = 0.5) +
      scale_fill_manual(values = colpalette_sex) +
      scale_color_manual(values = colpalette_sex) +
      scale_x_continuous(limits = c(4.5, 32)) +
      scale_y_continuous(limits = c(ymin, ymax),
                         breaks = y_breaks,
                         labels = scales::number_format(accuracy = 1)) +
      labs(title = unique(indiv_subdata$prot_id),
           subtitle = bquote(atop(
                      rho[F] == .(round(corr_subdata$estimate[corr_subdata$Sex == "Female"], 2)) * "," ~ 
                      p[adj,F] == .(formatC(corr_subdata$p_val_adj[corr_subdata$Sex == "Female"], format = "e", digits = 0)),
                      
                      rho[M] == .(round(corr_subdata$estimate[corr_subdata$Sex == "Male"], 2)) * "," ~ 
                      p[adj,M] == .(formatC(corr_subdata$p_val_adj[corr_subdata$Sex == "Male"], format = "e", digits = 0)))),
           x = "",
           y = "")
    
    if (manually_set_panel_tag == TRUE) {
    # Only runs when condition is TRUE
           corr_indiv <- (corr_indiv + labs(tag = panel_tags[k]) + theme(plot.tag = element_text(size = 25)))
            }
          
    # Store plot output in list      
    list_corr_prot[[k]] <- corr_indiv
  }

  # Return a list of the plots as the output
  return(list_corr_prot)
  # return(indiv_subdata)
}

```

## Fold-change Bar plots


```{r function_fc_barplot}

barplot_fc_magnet <- function(dataframe, filename_string){

      fc_barplot  <- (ggplot(data=dataframe, aes(x=reorder(Protein,Order), y=Log2FC, fill=type)) +

          geom_col(color="black", position=position_dodge2(width = 0.8, reverse = FALSE))+
        # Line range to show error
          geom_linerange(aes(x=reorder(Protein,Order), ymin=(Log2FC-FC_SD_Log2), ymax=(Log2FC+FC_SD_Log2)),
                      position=position_dodge(0.9))+
        theme_minimal()+
        theme(panel.grid.major.x = element_blank(),
              panel.grid.major.y = element_line(color="azure4"),
              panel.grid.minor.y = element_line(color="azure3"),
              axis.title=element_text(size=18,face="bold"),
              axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=18, color = "black"),
              axis.text.y = element_text(size=18, color = "black"),
              legend.position="right",
              legend.title = element_text(size = 18),
              legend.text = element_text(size = 16),
              legend.margin=margin(t = 0, r = 0.1, b = 0, l = 0, "in"),
              legend.key.size = unit(0.4, "in"),
              axis.title.x = element_blank())+
        geom_hline(yintercept=0, linetype="solid", color = "black", linewidth=2) +
        ylab(yaxval_edbar) +
        scale_fill_manual(name = "Status",values=c(Enriched="blue", Depleted="red2", "Not Significant"="grey40"))
          )
      
      # Export plot as svg
      ggsave(file = paste("rmd_output//plots//",basename,filename_string,".svg", sep=""), plot=fc_barplot, width=10, height=7)

      return(fc_barplot)
}

```



<br/><br/>


# Tidying {.tabset .tabset-fade}

$~$

## Tidying - Aging Cohort

```{r aging_cohort_dia_tidying}

#################################################################################
#################################################################################
## 
##      Metadata

# Cleaning up replicate name
    # Remove common prefix
    metadata$replicate <- sub("*X06_01_01_NeoV_PepMap5mm_PepSep8cmFossil_Astral_24min_Churchill_", "", metadata$replicate)
    
    # Remove run order designation
    metadata <- metadata %>%
                mutate(replicate = str_replace_all(replicate, "_\\d+$", ""))  
    
    
# Fixing single IEQC that was mistakenly left blank in the "Sex" column
    # metadata$Sex <- metadata$Sex %>% 
    #                 replace_na('Mixed')

    # Renaming Plate (batches) as "A" and "B" instead of 1 and 2
    metadata$Plate <- sub("1", "A", metadata$Plate)
    metadata$Plate <- sub("2", "B", metadata$Plate)                      
  

# Ensuring metadata is reordered by replicate to match to sample data later.
    metadata = metadata[order(metadata$replicate),]

    

#################################################################################
#################################################################################
## 
##      Peptide-level DIA data

## DATAFRAMES ##
    pep_unnorm_meta_l <- generate_long_dia_meta(dataframe = pep_unnorm, metadata_dataframe = metadata)
    pep_norm_meta_l <- generate_long_dia_meta(dataframe = pep_norm, metadata_dataframe = metadata)
    
    pep_unnorm_w <- tidy_wide_colnames(dataframe = pep_unnorm)
    pep_norm_w <- tidy_wide_colnames(dataframe = pep_norm)

## DATA MATRICES ##
    pep_unnorm_matrix <- generate_pep_matrix_dia(dataframe = pep_unnorm_w)
    pep_norm_matrix <- generate_pep_matrix_dia(dataframe = pep_norm_w)

#################################################################################
#################################################################################
## 
##      Protein-level DIA data

## DATAFRAMES ##
    prot_unnorm_meta_l <- generate_long_dia_meta(dataframe = prot_unnorm, metadata_dataframe = metadata)
    prot_norm_meta_l <- generate_long_dia_meta(dataframe = prot_norm, metadata_dataframe = metadata)
    
    prot_unnorm_w <- tidy_wide_colnames(dataframe = prot_unnorm)
    prot_norm_w <- tidy_wide_colnames(dataframe = prot_norm)

## DATA MATRICES ##
    prot_unnorm_matrix <- generate_prot_matrix_dia(dataframe = prot_unnorm_w)
    prot_norm_matrix <- generate_prot_matrix_dia(dataframe = prot_norm_w)


```

<br/><br/>

## Shared Color Palettes

```{r shared_aging_color_palettes}
# Setting color palettes for continuous variables to be plotted:
    viridis_colors <- viridis(9)
    # Age (in weeks or months)  
        colpalette_age = viridis_colors

    # Instrument run order
        colfunc_order <- colorRampPalette(c("magenta", "black"))
        colpalette_order <- colfunc_order(15)

# Setting color palettes for discrete variables to be plotted:
    # Sex
        colpalette_sex <- c("Female" = "#5D3A9B", "Male"= "#E66100", "Mixed"= "azure4")

    # Hemolysis status
        colpalette_hemolysis <- c("Very severe" = "red4", "Severe" = "red3", "Moderately severe" = "indianred1", "Moderate" = "salmon", "Moderately low" = "lightpink", "Low" = "pink", "Very low" = "beige")
    
    # Age group  
        colpalette_group <- c(setNames(viridis_colors, as.character(1:9)),  
                              "IEQC" = "#8B0000",
                              "IBQC" = "#C3C2C2")
        
# Plate or sample batch (1 = A and 2 = B)
        colpalette_plate <- c("A" = "#000000", "B"= "#40B0A6")

```

<br/><br/>

# Figure 1 {.tabset .tabset-fade}

The timeline component of Figure 1A was generated in R and assembled in Inkscape with the young and aged mouse, sex label, cohort table (Figure 1B), and experimental workflow summary (Figure 1C).

$~$

## Tidying

```{r fig1_tidy}
# Setting-up numbered dataframe from 1-31 to be paired with approximate age spacing in later timeline-based part of figure
background_bins <- data.frame("months" = seq(1, 31, by=1))

# Note that Group 8 in excel = 127-131, 
cohort_timeline <- data.frame("SampleGroup"=c(1,2,3,4,5,6,7,9,8), 
                              "Chronological"=c(1,2,3,4,5,6,7,8,9),
                              "months_label"= c("5.9","8.2","11.2","14.3","17.5","20.4","22.4","27","29.3-30.3"),
                              "months"= c(5.9,8.2,11.2,14.3,17.5,20.4,22.4,27,29.8), # Includes median of the two timepoints here
                              "weeks_label"=c("25","35","48","62","75","89","97","115-117","127-131"),
                              "weeks"=c(25,35,48,62,75,89,97,116,129), # Includes median of the two timepoints here
                              "weeks_label"=c("25","35","48","62","75","89","97","115-117","127-131"),
                              "position_x"=c(5.1,7.4,10.3,13.4,16.6,19.6,21.6,26.3,28.3),
                              "position_y"=c(0.3,0.3,0.3,0.3,0.3,0.3,0.3,0.3,0.3),
                              "orientation"=c(1,-1,1,-1,1,-1,1,-1,1))

```


## Figure 1A

```{r fig1a_gradient_timeline}

fig1_female_gradient <- fig1_gradients(dataframe = background_bins, 
                                  y_parameter = background_bins$months,
                                  fill_parameter = background_bins$months,
                                  median_lifespan = as.numeric(866 / (365/12)), # median lifespan in days, converted to months
                                  color_low = "#660066", 
                                  color_mid = "#CC9EFA",
                                  # color_mid = "#CC66FF",
                                  color_high = "#FAF0FF")

fig1_male_gradient <- fig1_gradients(dataframe = background_bins, 
                                  y_parameter = background_bins$months,
                                  fill_parameter = background_bins$months,
                                  median_lifespan = as.numeric(901 / (365/12)), # median lifespan in days, converted to months
                                  color_low = "#CC3300", 
                                  color_mid = "#FACC9E", 
                                  color_high = "#FFEEDD")

fig1_age_timeline <- (ggplot(data = cohort_timeline, aes(x=months, y=position_y)) +
          theme_void()+
          theme(plot.margin = margin(0, 0.2, 0, 0.2, "in"),  # t, r, b, l 
                panel.spacing = unit(0, "pt"),
                axis.title.x=element_blank(),
                axis.title.y=element_blank(),
                axis.text.x=element_blank(),
                axis.text.y=element_blank(),
                axis.ticks.x=element_blank(),
                axis.ticks.y=element_blank(),
                axis.line.x=element_blank(),
                axis.line.y=element_blank()) +
          coord_fixed(xlim=c(1,32),
                  ylim = c(0, 0.6),
                  clip = 'off') +   
          geom_segment(data=cohort_timeline, aes(y=0.3,yend=0.3,x=1,xend=32), color='black', size=0.5) +

          geom_segment(data=cohort_timeline, aes(y=position_y,yend=(position_y+(1.25*orientation)),x=months,xend=months), color='black', size=0.2) +
          geom_point(data = cohort_timeline, aes(x=months, y=position_y), size=1, shape=20, fill = "black") +

          geom_label(data=cohort_timeline, aes(x=position_x, y=(position_y+(1.25*orientation)),label=months_label),size=1.5,vjust=0.5,hjust= 0.05,color='black',label.size = NA))

```

```{r fig1a_partial}
 
fig1_timeline <- (fig1_female_gradient / fig1_age_timeline / fig1_male_gradient + plot_layout(heights = c(1,1.5,1)) + plot_annotation(tag_levels = list(c("A", "",  ""))) & theme(plot.tag = element_text(size = 5)))


ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"fig1a_timeline.svg", sep=""), plot=fig1_timeline, width=(2.5), height=(0.85))

```

```{r print_fig1a, fig.dim = c(2.5, 0.85)}

fig1_timeline

```


<br/><br/>


# Figure 2 & 3 {.tabset .tabset-fade}

__Figure 2:__ Evaluating age-related heterogeneity, top-line changes with age, and picking out some specific markers known to be related to aging. 

* Protein CV in aging comparing youngest and oldest mice.

* PCA plot of all proteins colored by aging by age of mouse

* PCA plot of proteins that correlate with age colored by age of mouse

* Heatmap of proteins that correlate with age

* 5 scatter plots of age, senescence, and frailty-related proteins printed with Spearman Rho and the p-value of the fitted linear model.
 
__Figure 3:__ Pulling out the top most increased and decreased proteins with age. Looking at Reactome pathways from proteins that correlated with age.

* Scatter plots of 5 proteins that increase most by Spearman rho with age printed with Spearman Rho and the p-value of the fitted linear model.
 
* Scatter plots of 5 proteins that decrease most by Spearman rho with age printed with Spearman Rho and the p-value of the fitted linear model.

* Reactome pathways sorted by the most statistically signficant after multiple hypothesis testing correction.

<br/><br/>

## Tidying

```{r fig2_tidy}

# Retooling dataframe for control evaluation
    prot_norm_meta_l$norm_cond <- "Median"
    prot_unnorm_meta_l$norm_cond <- "Raw"
    
    prot_full_meta_l <- rbind(prot_norm_meta_l, prot_unnorm_meta_l)
    
    prot_full_meta_l$norm_cond <- factor(prot_full_meta_l$norm_cond, levels=c("Raw", "Median"))

# Retooling dataframe for control evaluation
    pep_norm_meta_l$norm_cond <- "Median"
    pep_unnorm_meta_l$norm_cond <- "Raw"
    
    pep_full_meta_l <- rbind(pep_norm_meta_l, pep_unnorm_meta_l)
    
    pep_full_meta_l$norm_cond <- factor(pep_full_meta_l$norm_cond, levels=c("Raw", "Median"))

```

<br/><br/>

## CV Density

```{r fig2a}

fig2a <- plot_cv_dens(dataframe = (prot_norm_meta_l %>%
                                   filter(SampleGroup == "1" | SampleGroup == "9") %>%
                                   filter(protein != "sp|P00924|ENO1_YEAST")),
                      analyte = "protein",
                      title_string = "Protein CV in aging",
                      grouping_vars = c("protein","SampleGroup"),
                      colorvector <- c("1"="chocolate", "9"="dodgerblue"),
                      legtitle = "Age \n(Months)",
                      labels = c("1" = "5.85", "9" = "29-31"),
                      xaxmax = 200,     #x axis maximum limit
                      yaxmax = 0.02)    #y axis maximum limit    
fig2a


```

<br/><br/>

## PCA analysis - All proteins

```{r fig2b}

fig2b <- plot_pca_continuous(data_matrix_for_pca = prot_norm_matrix,
                           metadata_type_label = "Age\n(Months)",
                           metadata_column_color = metadata$Age_months,
                           color_palette = colpalette_age,
                           legend_title_lower_margin = 0.2)                        
fig2b <- fig2b + ggtitle("All proteins") + theme(plot.title = element_text(size = 18, hjust = 0.5))

ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"fig2b.svg", sep=""), plot=fig2b, width=7, height=6)   
ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"fig2b.png", sep=""), plot=fig2b, width=7, height=6)   

```
<br/><br/>

## Spearman Correlation - All

```{r spearman_corr_all}


corr_spear_prot <- corrspear_function_ageVSabundance(dataframe = (prot_norm_meta_l %>%
                                                                 group_by(., protein)),
                                                                 analyte_type ="protein")

corr_spear_prot_pval <- corrspear_filtering_pval(dataframe = corr_spear_prot,
                                                analyte_type = "protein",
                                                adjusted_p_value_cutoff = 0.05)     

corr_spear_prot_pval_rho  <- corrspear_filtering_pval_rho(dataframe = corr_spear_prot, 
                                                         analyte_type = "protein",
                                                         adjusted_p_value_cutoff = 0.05, 
                                                         lower_spearman_rho = -0.3, 
                                                         upper_spearman_rho = 0.3)


corr_spear_pep <- corrspear_function_ageVSabundance(dataframe = (pep_norm_meta_l %>%
                                                                     group_by(., modifiedSequence)),
                                                                     analyte_type ="peptide")

corr_spear_pep_pval <- corrspear_filtering_pval(dataframe = corr_spear_pep,
                                                analyte_type = "peptide",
                                                adjusted_p_value_cutoff = 0.05)     

corr_spear_pep_pval_rho  <- corrspear_filtering_pval_rho(dataframe = corr_spear_pep, 
                                                         analyte_type = "peptide",
                                                         adjusted_p_value_cutoff = 0.05, 
                                                         lower_spearman_rho = -0.3, 
                                                         upper_spearman_rho = 0.3)
 
```

```{r correlated_prot_df_matrix}
# Narrowing down input 
    corr_prots_all <- as.data.frame(unique(corr_spear_prot_pval_rho$protein))
    corr_prots_all <- corr_prots_all %>%
                      rename("protein" = "unique(corr_spear_prot_pval_rho$protein)")

# Filtering wide_format protein data based on the most significant proteins by spearman
    prot_corr_sig_df <- left_join(corr_prots_all, prot_norm_w, by = "protein")
    
# Converting filtered data to a matrix
    prot_corr_sig_matrix <- generate_prot_matrix_dia(dataframe = prot_corr_sig_df)

```


## PCA Analysis - Correlated with Age

```{r}
fig2c <- plot_pca_continuous(data_matrix_for_pca = prot_corr_sig_matrix,
                           metadata_type_label = "Age\n(Months)",
                           metadata_column_color = metadata$Age_months,
                           color_palette = colpalette_age,
                           legend_title_lower_margin = 0.2)      

fig2c <- fig2c + ggtitle("Proteins correlated with age") + theme(plot.title = element_text(size = 18, hjust = 0.5))


ggsave(file = paste(working_directory, "\\rmd_output\\plots\\",basename,filenamebase,"fig2c.svg", sep=""), plot=fig2c, width=7, height=6)   
ggsave(file = paste(working_directory, "\\rmd_output\\plots\\",basename,filenamebase,"fig2c.png", sep=""), plot=fig2c, width=7, height=6)   

```

## Heatmap -  Correlated with age

```{r heatmap_input_colorschemes}

# Define color scale for the heatmap
colpalette_heatmap <- colorRampPalette(c("navy", "white", "firebrick"))(20)

# Define a color palette for the annotation variables
    annotation_colors <- list(
                         Sex = c(Male = "#E66100", Female = "#5D3A9B"),
                         "Age (Months)" = viridis(15),
                         Plate = c("A" = "#000000", "B"= "#4682B4"))

# Remove accession column and external QC samples
      prot_corr_heatmap_matrix <- prot_corr_sig_matrix[,!colnames(prot_corr_sig_matrix) %in% c('first_accession', 'IBQC01', 'IBQC02', 'IBQC03', 'IBQC04', 'IBQC05', 'IBQC06', 'IBQC07', 'IBQC08', 'IBQC09', 'IEQC01', 'IEQC02', 'IEQC03', 'IEQC04', 'IEQC05', 'IEQC06', 'IEQC07', 'IEQC08', 'IEQC09')]

# Paring down metadata
    heatmap_meta <- metadata %>%
                    filter(SampleGroup != "IBQC" & SampleGroup != "IEQC") %>%
                    select(c(replicate, Sex, Age_months, Plate)) %>%
                    rename("Age (Months)" = "Age_months")

# Setting replicates as rownames
    rownames(heatmap_meta) <- heatmap_meta$replicate

# Optionally, remove the 'ID' column from the data frame
    heatmap_meta$replicate <- NULL
    
# Checking that metadata rows match matrix columns    
    all(rownames(heatmap_meta) == colnames(prot_corr_heatmap_matrix))  # Should return TRUE

```

```{r fig2d}
# Cluster the columns (samples)
col_dist <- dist(t(prot_corr_heatmap_matrix))
col_clust <- hclust(col_dist, method = "complete")

# Convert to dendrogram object
col_dend <- as.dendrogram(col_clust)

# Moving one cluster of samples to the other side.
    # sample order: keep current order but move the target samples to the end
      current_order <- labels(col_dend)
      target_samples <- c("JXC0018", "JXC0019", "JXC0026")
      non_target_samples <- setdiff(current_order, target_samples)
      new_order <- c(non_target_samples, target_samples)

    # Rotate the dendrogram
      rotated_col_dend <- rotate(col_dend, order = new_order)
      
      rotated_col_clust <- as.hclust(rotated_col_dend)
      
      new_sample_order <- rotated_col_clust$labels[rotated_col_clust$order]


# Reorder the metadata to match the new column order
    heatmap_meta_ordered <- heatmap_meta[new_sample_order, , drop = FALSE]

# Double-check metadata and sample data match
    stopifnot(identical(rownames(heatmap_meta_ordered), new_sample_order))
    all(rownames(heatmap_meta_ordered) == new_sample_order)  # Should return TRUE

# Fully-clustered heatmap
    fig2d_rotated <- pheatmap(prot_corr_heatmap_matrix,
                      color = colpalette_heatmap,
                      cluster_rows = TRUE,
                      cluster_cols = rotated_col_clust,
                      annotation_col = heatmap_meta_ordered,
                      annotation_colors = annotation_colors,
                      show_rownames = FALSE,
                      show_colnames = FALSE,
                      fontsize = 17,
                      breaks = seq(-4, 4, length.out = 20),
                      scale = "row",
                      main = "Proteins Correlated with Age")

# Saving heatmap in environment as a ggplot object to use in Patchwork later.
fig2d_rotated <- as.ggplot(fig2d_rotated)

ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"fig2d_rotated.jpg", sep=""), plot=fig2d_rotated, width=10, height=12)   

```


## Protein Correlation Plots

Setting-up vector of a subset of age-related proteins from literature to consider

```{r corr_aging_targets}

# 5 known age-related markers found in the dataset: "PAI1", "OSTP", "CALR", "GRN", and "ICAM1"
    corr_targets <- c("sp|P22777|PAI1_MOUSE", "sp|P10923|OSTP_MOUSE", "sp|P14211|CALR_MOUSE", "sp|P28798|GRN_MOUSE", "sp|P13597|ICAM1_MOUSE")
    
```


```{r corr_all_subdata}
# Get first uniprot IDs into column to make for easier labeling later.
    corr_prot_prho_inc <- corr_spear_prot_pval_rho %>%
                             filter(estimate > 0.3) %>%
                             get_1st_uniprot(.)                        
    
    corr_prot_prho_dec <- corr_spear_prot_pval_rho %>%
                         filter(estimate < -0.3) %>%
                         get_1st_uniprot(.)

# Extracting the individual normalized MS data for the above age-related markers
    corr_prot_prho_aging <- prot_norm_meta_l %>%
                            filter(SampleGroup != "IEQC" & SampleGroup != "IBQC") %>%
                            filter(protein != "sp|P00924|ENO1_YEAST") %>%
                            subset(., protein %in% corr_targets) 

# Assigning a label in new column "Trend" to assign plot colors later.  
    corr_prot_prho_aging$Trend <- "Aging and Senescence"
     
# Extracting the 5 most correlated proteins from the full dataset that have the most positively or the most negatively correlated proteins by Spearman correlation that also have a Benjamin-Hochberg corrected p-value below 0.05. Additionally, assigning a label in new column "Trend" to assign plot colors later.    
    corr_top5_inc <- extract_top5(dataframe_indiv = prot_norm_meta_l,
                     dataframe_corr = corr_prot_prho_inc, 
                     ordered_col = dplyr::desc(corr_prot_prho_inc$estimate),
                     trend_id = "Increases")

    corr_top5_dec <- extract_top5(dataframe_indiv = prot_norm_meta_l,
                     dataframe_corr = corr_prot_prho_dec, 
                     ordered_col = corr_prot_prho_dec$estimate,
                     trend_id = "Decreases")
# Generating a combined dataframe of the aging markers and the 5 most positively correlated and the 5 most negatively correlated proteins.    
    corr_fig2 <- rbind(corr_prot_prho_aging, corr_top5_inc, corr_top5_dec)

# Generating a vector listing the full protein names from the combined dataftame to iterate through for plot generation later..        
    target_proteins <- c(unique(corr_prot_prho_aging$protein), unique(corr_top5_inc$protein), unique(corr_top5_dec$protein))
    
```


```{r indiv_corr_lm_all}

# Define list for proteins of interest to save
    list_corr_prot_all <- list()
    custom_tags_all <- NA

    # Run function
    list_corr_prot_all <- generate_indiv_corr_plots(individual_data = corr_fig2, 
                                       correlation_data = corr_spear_prot,
                                       targets_corr = target_proteins, 
                                       list_corr_prot = list_corr_prot_all,
                                       manually_set_panel_tag = FALSE,
                                       panel_tags = custom_tags_all)
# , "", "C", "", "D", "", "E", "", "F", "", "G", ""
```


```{r indiv_corr_lm_fig2_fig3}

summary <- (((list_corr_prot_all[[1]] | list_corr_prot_all[[2]] | list_corr_prot_all[[3]] | list_corr_prot_all[[4]] | list_corr_prot_all[[5]]) / (list_corr_prot_all[[6]] | list_corr_prot_all[[7]] | list_corr_prot_all[[8]] | list_corr_prot_all[[9]] | list_corr_prot_all[[10]]) / (list_corr_prot_all[[11]] | list_corr_prot_all[[12]] | list_corr_prot_all[[13]] | list_corr_prot_all[[14]] | list_corr_prot_all[[15]])) + plot_layout(nrow = 3, heights = c(1:1))) #, guides = "collect"))

  ggsave(file = paste(working_directory, "\\rmd_output\\plots\\",basename,filenamebase,"corr_plot_all_sumary.png", sep=""), plot=summary, width=14, height=10)



  corr_y_axis_expr <- expression(Log[2]*"(Normalized Area)")
  
  
list_corr_prot_all[[1]] <- (list_corr_prot_all[[1]] + theme(axis.title.x = element_blank()) + ylab(corr_y_axis_expr))
list_corr_prot_all[[2]] <- (list_corr_prot_all[[2]] + theme(axis.title = element_blank())) 
list_corr_prot_all[[3]] <- (list_corr_prot_all[[3]] + theme(axis.title.y = element_blank()) + xlab("Age (Months)"))
list_corr_prot_all[[4]] <- (list_corr_prot_all[[4]] + theme(axis.title = element_blank()))
list_corr_prot_all[[5]] <- (list_corr_prot_all[[5]] + theme(axis.title = element_blank()))


fig2e <- as.ggplot(list_corr_prot_all[[1]] | list_corr_prot_all[[2]] | list_corr_prot_all[[3]] | list_corr_prot_all[[4]] | list_corr_prot_all[[5]])

  ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"fig2e.png", sep=""), plot=fig2e, width=3, height=12)

list_corr_prot_all[[6]] <- (list_corr_prot_all[[6]] + theme(axis.title.x = element_blank()) + ylab(corr_y_axis_expr))

list_corr_prot_all[[7]] <- (list_corr_prot_all[[7]] + theme(axis.title = element_blank())) 

list_corr_prot_all[[8]] <- (list_corr_prot_all[[8]] + theme(axis.title.y = element_blank()) + xlab("Age (Months)"))

list_corr_prot_all[[9]] <- (list_corr_prot_all[[9]] + theme(axis.title = element_blank()))

list_corr_prot_all[[10]] <- (list_corr_prot_all[[10]] + theme(axis.title = element_blank())) 


list_corr_prot_all[[11]] <- (list_corr_prot_all[[11]] + theme(axis.title.x = element_blank()) + ylab(corr_y_axis_expr))

list_corr_prot_all[[12]] <- (list_corr_prot_all[[12]] + theme(axis.title = element_blank()))

list_corr_prot_all[[13]] <- (list_corr_prot_all[[13]] + theme(axis.title.y = element_blank()) + xlab("Age (Months)"))

list_corr_prot_all[[14]] <- (list_corr_prot_all[[14]] + theme(axis.title = element_blank()))
                        
list_corr_prot_all[[15]] <- (list_corr_prot_all[[15]] + theme(axis.title = element_blank()))



fig3a <- as.ggplot((list_corr_prot_all[[6]] | list_corr_prot_all[[7]] | list_corr_prot_all[[8]] | list_corr_prot_all[[9]] | list_corr_prot_all[[10]]))
          
fig3b <- as.ggplot((list_corr_prot_all[[11]] | list_corr_prot_all[[12]] | list_corr_prot_all[[13]] | list_corr_prot_all[[14]] | list_corr_prot_all[[15]]))



ggsave(file = paste(working_directory, "\\rmd_output\\plots\\",basename,filenamebase,"fig3a.png", sep=""), plot=fig3a, width=12, height=5)

ggsave(file = paste(working_directory, "\\rmd_output\\plots\\",basename,filenamebase,"fig3b.png", sep=""), plot=fig3b, width=12, height=5)



```


<br/><br/>


## Reactome Pathways from Age Correlation

```{r fig3c}
# Reactome analysis of filtered proteins identified in the aging cohort that correlate positively with age and cross significant threshold
    prot_aging_incr_reactome <- gene_id_reactome_analysis(dataframe_and_Column = corr_prot_prho_inc$first_accession,
                                                          data_summary = "all_prot_age_increase")

# Reactome analysis of filtered proteins identified in the aging cohort that correlate negatively with age and cross significant threshold
    prot_aging_decr_reactome <- gene_id_reactome_analysis(dataframe_and_Column = corr_prot_prho_dec$first_accession,
                                                          data_summary = "all_prot_age_decrease")
    
# Extract top 5 entries from the returned dataframe    
 reactome_inc_df <- as.data.frame(prot_aging_incr_reactome) %>%
                    arrange(p.adjust) %>%
                    slice_head(n = 5) %>%
                    mutate(category = "Increase",
                           p_adj_log10 = -log10(p.adjust))
 
 
reactome_dec_df <- as.data.frame(prot_aging_decr_reactome) %>%
                   arrange(p.adjust) %>%
                   slice_head(n = 5) %>%
                   mutate(category = "Decrease",
                          p_adj_log10 = -log10(p.adjust))

# Combine increasing/decreasing Reactome dataframes
    reactome_plot_data <- bind_rows(reactome_inc_df, reactome_dec_df) %>%
                          mutate(category = factor(category, levels = c("Increase", "Decrease")))

# Sort the data within each category by its odds ratio  
      inc_order <- reactome_plot_data %>%
        filter(category == "Increase") %>%
        arrange(OddsRatio) %>%
        pull(Description)
    
    dec_order <- reactome_plot_data %>%
        filter(category == "Decrease") %>%
        arrange(desc(OddsRatio)) %>%
        # arrange(OddsRatio) %>%
        pull(Description)

# Increases plot on top (higher y), so their labels come last in factor levels
    reactome_plot_data_sort <- reactome_plot_data %>%
                                mutate(Description = factor(Description, 
                                       levels = c(dec_order, inc_order)))
  
# Set-up legend customization
  legend_or <- "Odds Ratio"

fig3c <- ggplot(reactome_plot_data_sort,
                aes(x = OddsRatio,
                    y = Description)) +
        
        # Setting-up increasing points (blue gradient)
            geom_point(data = subset(reactome_plot_data_sort, category == "Increase"),
                       aes(size  = Count, color = p_adj_log10), alpha = 0.9) +
            scale_color_gradientn(colors = c("lightskyblue", "dodgerblue", "navy"),
                                  name   = bquote(atop("Increases", -Log[10]~"(BH p-value)")),
                                  # name   = paste0("Increases\n", legend_pval),
                                  guide  = guide_colorbar(order = 1)) +
        
            new_scale_color() +
        
        # Setting-up decreasing points
            geom_point(data = subset(reactome_plot_data_sort, category == "Decrease"),
                       aes(size  = Count, color = p_adj_log10), alpha = 0.9) +
            scale_color_gradientn(colors = c("lightsalmon", "red", "firebrick"),
                                  name   = bquote(atop("Decreases", -Log[10]~"(BH p-value)")),
                                  # name   = paste0("Decreases\n", legend_pval),
                                  guide  = guide_colorbar(order = 2)) +
        
        # Setting a shared size scale
            scale_size_continuous(name = "Gene Count",
                                  range  = c(3, 10),  # min/max dot size in pts
                                  guide  = guide_legend(order = 3)) +
        
        # Vertical reference line at Odds ratio = 1 (no enrichment)
            geom_vline(xintercept = 1, linetype = "dashed", color = "grey50", linewidth = 0.4) +
        
        # Axes, scale, and theme
            labs(x   = "Odds Ratio",
                 y   = "Reactome Pathways",
                 tag = "C") +
            scale_y_discrete(expand = c(0.1, 0.5),
                             labels = function(y) str_wrap(y, width = 45)) +
            scale_x_continuous(expand = expansion(mult = c(0.05, 0.1))) +
            theme_minimal() +
            theme(plot.background  = element_blank(),
                  plot.margin      = margin(0, 0.2, 0, 0.2, "in"),
                  axis.text.y      = element_text(size = 12),
                  axis.text.x      = element_text(size = 13),
                  axis.title       = element_text(size = 14, face = "bold"),
                  legend.title     = element_text(size = 11),
                  legend.text      = element_text(size = 10),
                  legend.key.size  = unit(0.2, "in"),
                  legend.position  = "right",
                  legend.margin    = margin(0, 0, 0, 0))

ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"fig3c.svg", sep=""), plot=fig3c, width=(12), height=(8))
ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"fig3c.png", sep=""), plot=fig3c, width=(12), height=(8))
 
```


## Figure 2

Manually adjusted the positions of the heatmap annotation keys to ensure no text was cutoff and it was clear to read.


```{r fig2_assembly}

fig2a <- fig2a + theme(plot.margin = margin(t=0, r=0, b=0, l=0),
                       plot.title = element_text(margin = margin(t=0, r=0, b=6, l=0)))+ labs(tag = "A")
fig2b <- fig2b + theme(plot.margin = margin(t=0, r=0, b=0, l=4),
                       plot.title = element_text(margin = margin(t=0, r=0, b=6, l=0)))+ labs(tag = "B")
fig2c <- fig2c + theme(plot.margin = margin(t=5, r=0, b=0, l=5),
                       plot.title = element_text(margin = margin(t=0, r=0, b=6, l=0))) + labs(tag = "C")
fig2d_rotated <- fig2d_rotated + theme(plot.margin = margin(t=7, r=0, b=0, l=0))+ labs(tag = "D")
fig2e <- free(fig2e, side = "tbrl") + theme(plot.margin = margin(t=10, r=0, b=0, l=0))+ labs(tag = "E")

fig2_top_pca <- (((fig2b | fig2c) + plot_layout(ncol = 2, widths = c(1:1), guides = "collect")))

fig2_top_row <- ((fig2a | fig2_top_pca) + plot_layout(ncol = 2, widths = c(3.2, 6.8)))

fig2_partial <- ((fig2_top_row / free(fig2d_rotated, side = "rbl") / fig2e) + plot_layout(heights = c(1.5, 6, 2.75)) &
               theme(plot.margin = margin(5, 0, 0, 5), 
                     plot.tag = element_text(size = 25), 
                     plot.tag.position = c(0, 1)))

ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"fig2_partial.svg", sep=""), plot=fig2_partial, width=(12), height=(16))
# Moved Panel A to x = 50 and Panel B to 579.173

```

```{r print_fig2, fig.dim = c(15, 12)}

fig2_partial

```


## Figure 3

```{r fig2_and_fig3, results = 'hide'}

fig3a <- fig3a + theme(plot.margin = margin(t=0, r=0, b=0, l=0)) + labs(tag = "A")
fig3b <- fig3b + theme(plot.margin = margin(t=0, r=0, b=5, l=0)) + labs(tag = "B")
fig3c <- fig3c + theme(plot.margin = margin(t=5, r=0, b=0, l=0),
                       axis.text.y = element_text(size = 14)) +labs(tag = "C")


fig3 <- ((free(fig3a, side = "rl") / free(fig3b, side = "rl") / fig3c) + 
          plot_layout(nrow = 3, heights = c(3.5,3.5,5)) + guides(size = guide_legend(order = 1),
                                                                  fill = guide_colorbar(order = 2)) &
          theme(plot.margin = margin(5, 5 , 0 , 5), plot.tag = element_text(size = 25), plot.tag.position = c(0, 1)))

ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"fig3.png", sep=""), plot=fig3, width=(14), height=(12))


ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"fig3.svg", sep=""), plot=fig3, width=(14), height=(12))


```


```{r print_fig3, , fig.dim = c(10, 12)}

fig3

```


<br/><br/>


# Figure 4 {.tabset .tabset-fade}


Evaluating the contribution of sex to age-related protein changes by Spearman and OLS analysis.

$~$

Figure 4 generated in the python script by Mike Riffle (also on GitHub).



<br/><br/>



# Figure 5 {.tabset .tabset-fade}


Evaluating the protein-level Elastic Net model and OLS analyses.

$~$

Please refer to the Jupyter Notebook in GitHub generated by Mike Riffle for the Elastic Net analyses.

<br/><br/>




# Figure S1 {.tabset .tabset-fade}

Basic characterization of the inter-species nature of the Mag-Net method in mice.

$~$


## FC Calculations

Values obtained from Skyline's group comparison function at the protein-level using sum of transition areas (MS level = 2). The fold-changes, 95% Confidence intervals, and adjusted p-values were used as follows:

* Adjusted p-values transformed to Log~10~ values for volcano plot.

* Fold changes were transformed to Log~2~(fold change) values for volcano plot and barplots.

* The 95% Confidence intervals (Labeled as FC~max~ and FC~min~) were used to calculate standard error and standard deviation (SD) of the FC, and then the SD after Log~2~ transformation for Log~2~(fold change) plots.

$~$

Converting 95% CI to standard error:

  $$SE_{FC} = \frac{FC_{max} - FC_{min}}{\ 3.92}$$

Converting standard error to standard deviation:

  $$SD_{FC} = \frac{SE_{FC}}{\sqrt{\frac{1}{n_{EV}}+\frac{1}{n_{Plasma}}}}$$
  
Taking into account Log2 transformation on standard deviation:

  $$SD_{Log_{2}FC} = \frac{1}{ln(2)}*\frac{SD_{FC}}{FC}$$


## Tidying

```{r figS1_tidy}
#################################################################################
#################################################################################
## 
##      Figure S1A

# Remove MS Level and Fold Change Result column
Simplified_FC <- subset(magnet_dia, select = -c(MS.Level, Fold.Change.Result) ) 

# Numeric data including sample number for both groups and the number of standard error widths around the mean for a 95% confidence interval (A constant)
sample_n_SAX <- 3
sample_n_HYD <- 3
CI_SEwidth <- (2*1.96)

# Rename columns to simplify later steps, and use pipes add new columns with new values for plotting and later use.
Simplified_FC <- Simplified_FC %>%
                 rename("Gene" = "Protein.Gene",
                        "Group.Accessions" = "Protein.Accession",
                        "Group.Proteins" = "Protein",
                        "FC" = "Fold.Change",
                        "FCmin" = "Min.Fold.Change",
                        "FCmax" = "Max.Fold.Change",
                        "Adj_pValue" = "Adjusted.P.Value",
                        "Log2FC" = "Log.2.Fold.Change") %>%
                 mutate(Protein.Accession = Group.Accessions, # Duplicating group accession (Containing all accessions) to be split in next section
                 Log10pValue = log(Adj_pValue, 10), # Log 10 transform of adjusted p-value
                 Log2FCmin = log(FCmin, 2), # Log 2 transform of minimum fold change
                 Log2FCmax = log(FCmax, 2), # Log 2 transform of maximum fold change
                 FC_SE = ((FCmax-FCmin)/CI_SEwidth), # Converting 95% CI to standard error
                 FC_SD_NoProp = (FC_SE / (sqrt(1/sample_n_SAX)+(1/sample_n_HYD))), # Converting standard error to standard deviation
                 FC_SD_Log2 = ((1/log(2))*(FC_SD_NoProp/(FC)))) # Taking into account Log2 transformation on back-calculated standard deviation


# Simplifying the dataframe by selecting specific columns
    Simplified_FC <-  Simplified_FC[, c("Protein.Accession", "Group.Proteins", "Group.Accessions", "Gene", "FC", "FCmin", "FCmax", "Adj_pValue", "Log2FC", "Log10pValue", "Log2FCmin", "Log2FCmax", "FC_SE", "FC_SD_NoProp", "FC_SD_Log2")]

    # Generate new column for categories
    Simplified_FC$type <- "Not Significant"
    Simplified_FC$type[Simplified_FC$FC < 0.5 & Simplified_FC$Adj_pValue<0.05] <- "Depleted"
    Simplified_FC$type[Simplified_FC$FC > 2 & Simplified_FC$Adj_pValue<0.05] <- "Enriched"

# Combine dataframe of protein abundances with specific proteins of interest that will be highlighted later in barplots and volcano plot.
    Simplified_FC_splitGroups <- Simplified_FC %>%
                                 separate_rows(., Protein.Accession, sep = " / ", convert = FALSE)     
    
# 
    Simplified_FC_figs1 <- Simplified_FC_splitGroups %>%
                           left_join(magnet_prot_targets, ., by = "Protein.Accession") 
    

#################################################################################
#################################################################################
## 
##      Figure S1B

particle_count_A <- particle_count_A %>%
                    rename("Size" = "Bin.centre..nm.", 
                           "Average" = "Concentration.average",
                           "SD" = "Standard.Deviation") %>%
                    select(c(Size, Average, SD)) %>%
                    mutate(processing_replicate = "A")

particle_count_B <- particle_count_B %>%
                    rename("Size" = "Bin.centre..nm.", 
                           "Average" = "Concentration.average",
                           "SD" = "Standard.Deviation") %>%
                    select(c(Size, Average, SD)) %>%
                    mutate(processing_replicate = "B")

particle_counts <- rbind(particle_count_A, particle_count_B)

```


</br></br>

## Figure S1

```{r figS1_plotting}

#################################################################################
#################################################################################
##
##      Figure S1A and S1B

yaxval_edbar <- expression("Log"[2]~"(EV fraction / Total plasma)")

figS1a <- barplot_fc_magnet(dataframe = Simplified_FC_figs1, 
                      filename_string = "figS1ab")
#################################################################################
#################################################################################
##
##      Figure S1b

# Plotting mean particle concentration of two technical replicate captures of the IEQC sample pool
figS1b <- (ggplot(particle_counts, aes(x=Size, y=Average, group=processing_replicate, color = processing_replicate, fill = processing_replicate))+
           geom_ribbon(aes(y=Average, ymin=Average-SD, ymax=Average+SD), alpha=.4, colour = NA, show.legend = FALSE) +
           geom_line(linewidth=1.5)+
           theme_minimal() +
           theme(plot.margin = margin(0.1, 0.1, 0.1, 0.1, "in"),  # t, r, b, l
                 panel.grid.major.x = element_line(color="azure3"),
                 panel.grid.major.y = element_line(color="azure3"),
                 panel.grid.minor.x = element_blank(),
                 panel.grid.minor.y = element_blank(),
                 axis.text.x = element_text(size=16, color = "black",angle = 90, vjust = 0.5, hjust=1),
                 axis.text.y = element_text(size=16, color = "black"),
                 axis.title.x = element_text(size = 18, color = "black", margin = margin(t = 0.15, r = 0, b = 0, l = 0, "in")),
                 axis.title.y = element_text(size = 18, color = "black", margin = margin(t = 0, r = 0.15, b = 0, l = 0, "in")),
                 legend.title = element_text(size = 18),
                 legend.text = element_text(size = 16),
                 legend.margin=margin(t = 0, r = 0.1, b = 0, l = 0, "in"),
                 legend.key.size = unit(0.4, "in")) +
           geom_hline(yintercept=0, linetype="solid", color = "black", linewidth=1) +
           geom_vline(xintercept=0, linetype="solid", color = "black", linewidth=1) +
           scale_y_continuous(name="Mean Particle Concentration/mL") +
           scale_x_continuous(name="Particle Size (nm)", limits=c(0, 400)) +
           scale_color_manual(name = "Technical\nReplicate", values = c("A" = "#994F00", "B" = "#006CD1")) +
           scale_fill_manual(values = c("A" = "#994F00", "B" = "#006CD1")))
    ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"figS1b.svg", sep=""), plot=figS1b, width=(10), height=(6))


```

```{r figS1}
# Full figure
figS1 <- ((figS1a / figS1b) + plot_layout(nrow  = 2, widths = c(1, 1))  + 
            plot_annotation(tag_levels = list(c("A", "B"))) & theme(plot.tag = element_text(size = 5)))

ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"figS1.svg", sep=""), plot=figS1, width=(10), height=(14))

ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"figS1.png", sep=""), plot=figS1, width=(10), height=(14))

```


```{r print_figS1, fig.dim = c(10, 14)}

figS1

```


<br/><br/>


# Figure S2 {.tabset .tabset-fade}


Supplementary Figure 2 plotting:

* Peptide CV of IB-QC

* Protein CV of IB-QC

* Peptide CV of IE-QC

* Protein CV of IE-QC

* Peptide PCA plot colored by LC-MS run order

* Protein PCA plot colored by LC-MS run order

* Peptide PCA plot colored by sample preparation batch

* Protein PCA plot colored by sample preparation batch

* Peptide PCA plot colored by qualitative hemolysis level

* Protein PCA plot colored by qualitative hemolysis level

* Peptide PCA plot colored by age group

* Protein PCA plot colored by age group

$~$

## Tidying

```{r figS2_tidy}

    # Extract median external qc sample CV of peptide-level
       cv_pep_extqc_samples <- ((pep_full_meta_l %>%
                                filter(protein != "sp|P00924|ENO1_YEAST") %>%
                                filter(SampleGroup == "IEQC" | SampleGroup == "IBQC")) %>%
                                select(c(modifiedSequence, SampleGroup, log2bc_area, norm_cond)) %>%
                                group_by(norm_cond, SampleGroup, modifiedSequence) %>%
                                summarise(mean_area = mean(remove_log2(log2bc_area)),
                                         sd_area = sd(remove_log2(log2bc_area)),
                                         cv = ((sd_area / mean_area)*100),
                                         .groups = 'drop') %>%
                                group_by(norm_cond, SampleGroup) %>%
                                summarise(mean_cv = mean(cv),
                                          median_cv = median(cv),
                                          .groups = 'drop'))

    # Extract median external qc sample CV of protein-level
       cv_prot_extqc_samples <- ((prot_full_meta_l %>%
                                filter(protein != "sp|P00924|ENO1_YEAST") %>%
                                filter(SampleGroup == "IEQC" | SampleGroup == "IBQC")) %>%
                                select(c(protein, SampleGroup, log2bc_area, norm_cond)) %>%
                                group_by(norm_cond, SampleGroup, protein) %>%
                                summarise(mean_area = mean(remove_log2(log2bc_area)),
                                         sd_area = sd(remove_log2(log2bc_area)),
                                         cv = ((sd_area / mean_area)*100),
                                         .groups = 'drop') %>%
                                group_by(norm_cond, SampleGroup) %>%
                                summarise(mean_cv = mean(cv),
                                          median_cv = median(cv),
                                          .groups = 'drop'))

```

## CV summary - External QCs

```{r aging_cv_ext_qc_samples}

# Extracting the the percent CV of the PEPTIDES found in the external QC samples run with the aging cohort samples 
    # Inter-experiment external QC sample
    pep_med_ieqc_raw <- cv_pep_extqc_samples$median_cv[cv_pep_extqc_samples$norm_cond == "Raw" & cv_pep_extqc_samples$SampleGroup == "IEQC"]  
    pep_med_ieqc_med <- cv_pep_extqc_samples$median_cv[cv_pep_extqc_samples$norm_cond == "Median" & cv_pep_extqc_samples$SampleGroup == "IEQC"]  
    
    # Inter-batch external QC sample (IBQC)   
    pep_med_ibqc_raw <- cv_pep_extqc_samples$median_cv[cv_pep_extqc_samples$norm_cond == "Raw" & cv_pep_extqc_samples$SampleGroup == "IBQC"]  
    pep_med_ibqc_med <- cv_pep_extqc_samples$median_cv[cv_pep_extqc_samples$norm_cond == "Median" & cv_pep_extqc_samples$SampleGroup == "IBQC"]  
    
    
    print(paste0("External sample QC Peptide CVs - IEQC Raw = ", pep_med_ieqc_raw, " and Median = ", pep_med_ieqc_med, "; IBQC Raw = ", pep_med_ibqc_raw, " and Median = ", pep_med_ibqc_med,"(Log2 transormation removed)"))
    

# Extracting the the percent CV of the PROTEINS found in the external QC samples run with the aging cohort samples 
    # Inter-experiment external QC sample
    prot_med_ieqc_raw <- cv_prot_extqc_samples$median_cv[cv_prot_extqc_samples$norm_cond == "Raw" & cv_prot_extqc_samples$SampleGroup == "IEQC"]  
    prot_med_ieqc_med <- cv_prot_extqc_samples$median_cv[cv_prot_extqc_samples$norm_cond == "Median" & cv_prot_extqc_samples$SampleGroup == "IEQC"]  
    
# Inter-batch external QC sample (IBQC)        
    prot_med_ibqc_raw <- cv_prot_extqc_samples$median_cv[cv_prot_extqc_samples$norm_cond == "Raw" & cv_prot_extqc_samples$SampleGroup == "IBQC"]  
    prot_med_ibqc_med <- cv_prot_extqc_samples$median_cv[cv_prot_extqc_samples$norm_cond == "Median" & cv_prot_extqc_samples$SampleGroup == "IBQC"]  
    
    
    print(paste0("External sample QC CVs - IEQC Raw = ", prot_med_ieqc_raw, " and Median = ", prot_med_ieqc_med, "; IBQC Raw = ", prot_med_ibqc_raw, " and Median = ", prot_med_ibqc_med,"(Log2 transormation removed)"))
    
    
    
```

## CV summary - Enolase

```{r dataset_cvs_enolase}

  
# Extracting the protein-level percent CV of the yeast enolase internal QC spiked into all of the external QC samples
       cv_calc_eno_extqc <- ((prot_full_meta_l %>%
                              filter(protein == "sp|P00924|ENO1_YEAST") %>%
                              filter(SampleGroup == "IEQC" | SampleGroup == "IBQC")) %>%
                              select(c(protein, SampleGroup, log2bc_area, norm_cond)) %>%
                              group_by(norm_cond,SampleGroup) %>%
                              summarise(mean_area = mean(remove_log2(log2bc_area)),
                                       sd_area = sd(remove_log2(log2bc_area)),
                                       cv = ((sd_area / mean_area)*100),
                                       .groups = 'drop'))
     
    eno_prot_ibqc_raw <-  cv_calc_eno_extqc$cv[cv_calc_eno_extqc$norm_cond == "Raw" & cv_calc_eno_extqc$SampleGroup == "IBQC"]  
    eno_prot_ibqc_median <-  cv_calc_eno_extqc$cv[cv_calc_eno_extqc$norm_cond == "Median" & cv_calc_eno_extqc$SampleGroup == "IBQC"]  
    
    eno_prot_ieqc_raw <-  cv_calc_eno_extqc$cv[cv_calc_eno_extqc$norm_cond == "Raw" & cv_calc_eno_extqc$SampleGroup == "IEQC"]  
    eno_prot_ieqc_median <-  cv_calc_eno_extqc$cv[cv_calc_eno_extqc$norm_cond == "Median" & cv_calc_eno_extqc$SampleGroup == "IEQC"] 
    
    print(paste0("External QC sample's Internal QC (enolase) CVs - IBQC: Raw data = ", eno_prot_ibqc_raw, " and Median = ",eno_prot_ibqc_median, "; and IEQC:  Raw data = ", eno_prot_ieqc_raw, " and Median = ",eno_prot_ieqc_median, " (Log2 transormation removed)"))

    
  
# Extracting the protein-level percent CV of the yeast enolase internal QC spiked into all of the aging cohort samples 
        cv_calc_eno_samples <- ((prot_full_meta_l %>%
                                filter(protein == "sp|P00924|ENO1_YEAST") %>%
                                       filter(SampleGroup != "IEQC") %>%
                                       filter(SampleGroup != "IBQC")) %>%
                                select(c(protein, SampleGroup, log2bc_area, norm_cond)) %>%
                                group_by(norm_cond) %>%
                                summarise(mean_area = mean(remove_log2(log2bc_area)),
                                         sd_area = sd(remove_log2(log2bc_area)),
                                         cv = ((sd_area / mean_area)*100),
                                         .groups = 'drop'))
     
    eno_prot_sample_raw <- cv_calc_eno_samples$cv[cv_calc_eno_samples$norm_cond == "Raw"]  
    eno_prot_sample_median <- cv_calc_eno_samples$cv[cv_calc_eno_samples$norm_cond == "Median"]
  
    print(paste0("Sample internal protein QC (enolase) CVs - Raw data = ", eno_prot_sample_raw, " and Median = ",eno_prot_sample_median, "(Log2 transormation removed)"))


```


## Figure S2

Evaluating the inter-batch and inter-experiment QC sample CVs and topline characterization of the dataset pre- and post-median normalization. Note that no batch correction was employed.

$~$

```{r figS2_plots_individual}

figS2pep_ibqc <- plot_cv_dens_qc(dataframe = (pep_full_meta_l %>%
                                           filter(protein != "sp|P00924|ENO1_YEAST") %>%
                                           filter(SampleGroup == "IBQC")),
                              analyte = "modifiedSequence",
                              title_string = "IB-QC: Peptide CV",
                              grouping_vars = c("norm_cond", "modifiedSequence"),
                              colorvector <- c("Raw"="darkgray", "Median"="salmon"),
                              legtitle = "Normalization",
                              labels = c("Raw", "Median"),
                              xaxmax = 100,     #x axis maximum limit
                              yaxmax = 0.04)  #y axis maximum limit    
                              
figS2pep_ibqc <- figS2pep_ibqc + theme(axis.title.x = element_blank())

figS2pep_ieqc <- plot_cv_dens_qc(dataframe = (pep_full_meta_l %>%
                                              filter(protein != "sp|P00924|ENO1_YEAST") %>%
                                              filter(SampleGroup == "IEQC")),
                                  analyte = "modifiedSequence",
                                  title_string = "IE-QC: Peptide CV",
                                  grouping_vars = c("norm_cond", "modifiedSequence"),
                                  colorvector <- c("Raw"="darkgray", "Median"="salmon"),
                                  legtitle = "Normalization",
                                  labels = c("Raw", "Median"),
                                  xaxmax = 100,     #x axis maximum limit
                                  yaxmax = 0.04)    #y axis maximum limit    

figS2pep_ieqc <- figS2pep_ieqc + theme(axis.title.x = element_blank(),
                                       axis.title.y = element_blank())

  

figS2prot_ibqc <- plot_cv_dens_qc(dataframe = (prot_full_meta_l %>%
                                           filter(protein != "sp|P00924|ENO1_YEAST") %>%
                                           filter(SampleGroup == "IBQC")),
                              analyte = "protein",
                              title_string = "IB-QC: Protein CV",
                              grouping_vars = c("norm_cond", "protein"),
                              colorvector <- c("Raw"="darkgray", "Median"="salmon"),
                              legtitle = "Normalization",
                              labels = c("Raw", "Median"),
                              xaxmax = 100,     #x axis maximum limit
                              yaxmax = 0.04)  #y axis maximum limit    
                              
  figS2prot_ieqc <- plot_cv_dens_qc(dataframe = (prot_full_meta_l %>%
                                              filter(protein != "sp|P00924|ENO1_YEAST") %>%
                                              filter(SampleGroup == "IEQC")),
                                  analyte = "protein",
                                  title_string = "IE-QC: Protein CV",
                                  grouping_vars = c("norm_cond", "protein"),
                                  colorvector <- c("Raw"="darkgray", "Median"="salmon"),
                                  legtitle = "Normalization",
                                  labels = c("Raw", "Median"),
                                  xaxmax = 100,     #x axis maximum limit
                                  yaxmax = 0.04)    #y axis maximum limit    
  figS2prot_ieqc <- figS2prot_ieqc + theme(axis.title.y = element_blank())
  
    
```

```{r figS2_cv_dens_patchwork}
figS2_partial <- (((figS2pep_ibqc | figS2pep_ieqc | figS2prot_ibqc | figS2prot_ieqc)) + plot_layout(nrow = 2, guides = "collect") + plot_annotation(tag_levels = list(c("A", "B", "C", "D")) & theme(plot.tag = element_text(size = 25))))

```



<br/><br/>


## Individual PCA plots

```{r qc_pca_indiv_plots}

figS2_pep_acq_raw <- (plot_pca_continuous(data_matrix_for_pca = pep_unnorm_matrix,
                           metadata_type_label = "Run Order",
                           metadata_column_color = metadata$Run.Order,
                           color_palette = colpalette_order,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank(),
                                                                     legend.position = "none",
                                                                     plot.title = element_text(size = 24, hjust = 1, margin = margin(b = 16))) + labs(title = "Sample Peptides")

figS2_pep_acq_norm <- (plot_pca_continuous(data_matrix_for_pca = pep_norm_matrix,
                           metadata_type_label = "Run Order",
                           metadata_column_color = metadata$Run.Order,
                           color_palette = colpalette_order,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank(),
                                                                     legend.position = "none")


figS2_prot_acq_raw <- (plot_pca_continuous(data_matrix_for_pca = prot_unnorm_matrix,
                           metadata_type_label = "Run Order",
                           metadata_column_color = metadata$Run.Order,
                           color_palette = colpalette_order,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank(),
                                                                     legend.position = "none",
                                                                     plot.title = element_text(size = 24, hjust = 0, margin = margin(b = 16))) + labs(title = "Sample Proteins") 

figS2_prot_acq_norm <- (plot_pca_continuous(data_matrix_for_pca = prot_norm_matrix,
                           metadata_type_label = "Run Order",
                           metadata_column_color = metadata$Run.Order,
                           color_palette = colpalette_order,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank())


###################################################################################################


figS2_pep_batch_raw <- (plot_pca_discrete(data_matrix_for_pca = pep_unnorm_matrix,
                           metadata_type_label = "Batch",
                           metadata_column_color = metadata$Plate,
                           color_palette = colpalette_plate,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank(),
                                                                     legend.position = "none")             

figS2_pep_batch_norm <- (plot_pca_discrete(data_matrix_for_pca = pep_norm_matrix,
                           metadata_type_label = "Batch",
                           metadata_column_color = metadata$Plate,
                           color_palette = colpalette_plate,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank(),
                                                                     legend.position = "none")

figS2_prot_batch_raw <- (plot_pca_discrete(data_matrix_for_pca = prot_unnorm_matrix,
                           metadata_type_label = "Batch",
                           metadata_column_color = metadata$Plate,
                           color_palette = colpalette_plate,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank(),
                                                                     legend.position = "none")             

figS2_prot_batch_norm <- (plot_pca_discrete(data_matrix_for_pca = prot_norm_matrix,
                           metadata_type_label = "Batch",
                           metadata_column_color = metadata$Plate,
                           color_palette = colpalette_plate,
                           legend_title_lower_margin = 0.2)) + theme(axis.text.x = element_blank(),
                                                                     axis.title.x = element_blank())


###################################################################################################


figS2_pep_hemolysis_raw <- (plot_pca_discrete(data_matrix_for_pca = pep_unnorm_matrix,
                           metadata_type_label = "Hemolysis",
                           metadata_column_color = metadata$Hemolysis,
                           color_palette = colpalette_hemolysis,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank(),
                                                                     legend.position = "none")             

figS2_pep_hemolysis_norm <- (plot_pca_discrete(data_matrix_for_pca = pep_norm_matrix,
                           metadata_type_label = "Hemolysis",
                           metadata_column_color = metadata$Hemolysis,
                           color_palette = colpalette_hemolysis,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank(),
                                                                     legend.position = "none")


figS2_prot_hemolysis_raw <- (plot_pca_discrete(data_matrix_for_pca = prot_unnorm_matrix,
                           metadata_type_label = "Hemolysis",
                           metadata_column_color = metadata$Hemolysis,
                           color_palette = colpalette_hemolysis,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank(),
                                                                     legend.position = "none")             

figS2_prot_hemolysis_norm <- (plot_pca_discrete(data_matrix_for_pca = prot_norm_matrix,
                           metadata_type_label = "Hemolysis",
                           metadata_column_color = metadata$Hemolysis,
                           color_palette = colpalette_hemolysis,
                           legend_title_lower_margin = 0.2)) + theme(axis.title.x = element_blank(),
                                                                     axis.text.x = element_blank()) +
                           scale_color_manual(values = colpalette_hemolysis,
                                              breaks = c("Very severe", "Severe", "Moderately severe", "Moderate", "Moderately low", "Low", "Very low")) + 
                           labs(color = "Hemolysis")  # Need to restate because reordering legend breaks.
                          

###################################################################################################


figS2_pep_group_raw <- (plot_pca_discrete(data_matrix_for_pca = pep_unnorm_matrix,
                           metadata_type_label = "Sample Group",
                           metadata_column_color = metadata$SampleGroup,
                           color_palette = colpalette_group,
                           legend_title_lower_margin = 0.2)) + theme(legend.position = "none")

figS2_pep_group_norm <- (plot_pca_discrete(data_matrix_for_pca = pep_norm_matrix,
                           metadata_type_label = "Sample Group",
                           metadata_column_color = metadata$SampleGroup,
                           color_palette = colpalette_group,
                           legend_title_lower_margin = 0.2)) + theme(legend.position = "none")
                           
figS2_prot_group_raw <- (plot_pca_discrete(data_matrix_for_pca = prot_unnorm_matrix,
                           metadata_type_label = "Sample Group",
                           metadata_column_color = metadata$SampleGroup,
                           color_palette = colpalette_group,
                           legend_title_lower_margin = 0.2)) + theme(legend.position = "none")

figS2_prot_group_norm <- (plot_pca_discrete(data_matrix_for_pca = prot_norm_matrix,
                           metadata_type_label = "Sample Group",
                           metadata_column_color = metadata$SampleGroup,
                           color_palette = colpalette_group,
                           legend_title_lower_margin = 0.2))

figS2_prot_group_norm <- figS2_prot_group_norm + guides(fill = guide_legend(ncol = 2, title = "Age Group"))

```

## Figure assembly

Externally, the legends were repositioned in Inkscape for clearer visualization and exported as "figS2"

```{r figS2_patchwork}
figS2 <- (((figS2pep_ibqc | figS2prot_ibqc) / (figS2pep_ieqc | figS2prot_ieqc) / 
           (figS2_pep_acq_raw | figS2_pep_acq_norm | figS2_prot_acq_raw | figS2_prot_acq_norm) /
           (figS2_pep_batch_raw | figS2_pep_batch_norm | figS2_prot_batch_raw | figS2_prot_batch_norm) /
           (figS2_pep_hemolysis_raw | figS2_pep_hemolysis_norm | figS2_prot_hemolysis_raw | figS2_prot_hemolysis_norm) / 
           (figS2_pep_group_raw | figS2_pep_group_norm | figS2_prot_group_raw | figS2_prot_group_norm)) +
  plot_annotation(tag_levels = list(c("A", "B", "C", "D", "E", "", "F", "", "G", "", "H", "", "I", "", "J", "", "K", "", "L", ""))) +
  plot_layout(nrow = 6, guides = "collect") & theme(plot.tag = element_text(size = 25),
                                plot.tag.position = c(0, 1)))

# Saving plot as an .svg format.
    ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"figS2_partial.svg", sep=""), plot=figS2, width=16, height=18)    



```


## Figure S2

```{r print_figS2, fig.dim = c(16, 12)}

figS2

```

```{r subplot_removal}

# Removing individual plots to reduce clutter in environment
  rm(figS2pep_ibqc, figS2pep_ieqc, figS2prot_ibqc, figS2prot_ieqc, figS2_pep_group_raw, figS2_pep_group_norm, figS2_prot_group_raw, figS2_prot_group_norm, figS2_pep_acq_raw, figS2_pep_acq_norm, figS2_prot_acq_raw, figS2_prot_acq_norm, figS2_pep_batch_raw, figS2_pep_batch_norm, figS2_prot_batch_raw, figS2_prot_batch_norm, figS2_pep_hemolysis_raw, figS2_pep_hemolysis_norm, figS2_prot_hemolysis_raw, figS2_prot_hemolysis_norm)
```



<br/><br/>



# Figure S3 {.tabset .tabset-fade}

Supplementary Figure 3 combined a quick plot for peptide CV at the peptide-level between youngest and oldest mice with the Reactome Analysis sorted by the Odds Ratio (and passed significant after multiple hypothesis correction).

* Peptide CV in aging comparing youngest and oldest mice.

* Top 15 pathways associated with proteins that increased with age across full cohort.

* Top 15 pathways associated with proteins that decreased with age across full cohort.

## CV density

```{r figS3a}
figS3a <- plot_cv_dens(dataframe = (pep_norm_meta_l %>%
                                   filter(SampleGroup == "1" | SampleGroup == "9") %>%
                                   filter(protein != "sp|P00924|ENO1_YEAST")),
                      analyte = "modifiedSequence",
                      title_string = "Peptide CV in aging",
                      grouping_vars = c("modifiedSequence","SampleGroup"),
                      colorvector <- c("1"="chocolate", "9"="dodgerblue"),
                      legtitle = "Age \n(Months)",
                      labels = c("1" = "5.85", "9" = "29-31"),
                      xaxmax = 200,     #x axis maximum limit
                      yaxmax = 0.02) +
                      labs(tag = "A")
```

## Reactome by Odds Ratios

```{r figS3b_S3c}

# Extract top 15 by OddsRatio (already filtered to p.adjust < 0.05 by enrichPathway)
    reactome_inc_df_15 <- as.data.frame(prot_aging_incr_reactome) %>%
        arrange(desc(OddsRatio)) %>%
        slice_head(n = 15) %>%
        mutate(p_adj_log10 = -log10(p.adjust))
    
    reactome_dec_df_15 <- as.data.frame(prot_aging_decr_reactome) %>%
        arrange(desc(OddsRatio)) %>%
        slice_head(n = 15) %>%
        mutate(p_adj_log10 = -log10(p.adjust))


# Setting factor level ordering: ascending odds ratio so highest is at the top of the plot
    inc_levels_15 <- reactome_inc_df_15 %>%
        arrange(OddsRatio) %>%
        pull(Description)
    
    dec_levels_15 <- reactome_dec_df_15 %>%
        arrange(OddsRatio) %>%
        pull(Description)
    
    reactome_inc_df_15 <- reactome_inc_df_15 %>%
        mutate(Description = factor(Description, levels = inc_levels_15))
    
    reactome_dec_df_15 <- reactome_dec_df_15 %>%
        mutate(Description = factor(Description, levels = dec_levels_15))


# Setting shared axis limits
    or_max <- max(c(reactome_inc_df$OddsRatio, reactome_dec_df$OddsRatio), 
                  na.rm = TRUE) * 1.05

    pval_range_inc_15 <- range(reactome_inc_df_15$p_adj_log10, na.rm = TRUE)
    pval_range_dec_15 <- range(reactome_dec_df_15$p_adj_log10, na.rm = TRUE)


# Setting-up a shared theme
reactome_theme <- theme_minimal() +
                  theme(plot.background = element_blank(),
                        plot.margin     = margin(0.1, 0.3, 0.1, 0.1, "in"),
                        axis.text.y     = element_text(size = 11),
                        axis.text.x     = element_text(size = 12),
                        axis.title      = element_text(size = 13, face = "bold"),
                        legend.title    = element_text(size = 11),
                        legend.text     = element_text(size = 10),
                        legend.key.size = unit(0.2, "in"),
                        legend.position = "right",
                        legend.margin   = margin(0, 0, 0, 0))


# Proteins that increases with age
figS3b <- ggplot(reactome_inc_df_15,
                 aes(x = OddsRatio, y = Description)) +
                  geom_point(aes(size = Count, color = p_adj_log10), alpha = 0.9) +
                  scale_color_gradientn(colors = c("lightskyblue", "dodgerblue", "navy"),
                                        name   = bquote(atop("Increases", -Log[10]~"(BH p-value)")),
                                        limits = pval_range_inc_15,
                                        guide  = guide_colorbar(order = 1)) +
                  scale_size_continuous(name  = "Gene Count",
                                        range = c(3, 10),
                                        guide = guide_legend(order = 2)) +
                  geom_vline(xintercept = 1, linetype = "dashed",
                             color = "grey50", linewidth = 0.4) +
                  scale_y_discrete(expand = c(0.05, 0.5),
                                   labels = function(y) str_wrap(y, width = 40)) +
                  scale_x_continuous(limits = c(0, or_max),
                                     expand = expansion(mult = c(0.02, 0.05))) +
                  labs(x   = "Odds Ratio",
                       y   = "Reactome Pathways",
                       tag = "B") +
                  reactome_theme

# Proteins that decreases with age
figS3c <- ggplot(reactome_dec_df_15,
                 aes(x = OddsRatio, y = Description)) +
                  geom_point(aes(size = Count, color = p_adj_log10), alpha = 0.9) +
                  scale_color_gradientn(colors = c("lightsalmon", "red", "firebrick"),
                                        name   = bquote(atop("Decreases", -Log[10]~"(BH p-value)")),
                                        limits = pval_range_dec_15,
                                        guide  = guide_colorbar(order = 1)) +
                  scale_size_continuous(name  = "Gene Count",
                                        range = c(3, 10),
                                        guide = guide_legend(order = 2)) +
                  geom_vline(xintercept = 1, linetype = "dashed",
                             color = "grey50", linewidth = 0.4) +
                  scale_y_discrete(expand = c(0.05, 0.5),
                                   labels = function(y) str_wrap(y, width = 40)) +
                  scale_x_continuous(limits = c(0, or_max),
                                     expand = expansion(mult = c(0.02, 0.05))) +
                  labs(x   = "Odds Ratio",
                       y   = "Reactome Pathways",
                       tag = "C") +
                  reactome_theme
```

## Figure S3

```{r figS3}

figS3 <- ( figS3a + (figS3b | figS3c) + plot_layout(heights = c(0.75, 5), guides = "collect") & theme(plot.tag = element_text(size = 25), plot.tag.position = c(0, 1)))
  
ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"figS3.svg", sep=""), plot=figS3, width=(15), height=(14))

ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"figS3.png", sep=""), plot=figS3, width=(15), height=(14))

```


```{r print_figS3, fig.dim = c(15, 14)}

figS3

```



<br/><br/>

# Figure S4 {.tabset .tabset-fade}

Supplementary Figure 4 summarizes sex-related panels including PCA plots colored by sex and pulling out the 5 age-related markers that were shown in Figure 2 for the full cohort evaluated together:

* PCA analysis of all peptides colored by sex 

* PCA analysis of all proteins colored by sex

* PCA analysis of proteins correlated with age, colored by sex

* Scatter plot of proteins split by sex printed with Spearman Rho and the p-value of the fitted linear model.

$~$

## Spearman Correlation - By Sex

Females

```{r spearman_corr_females}

# Females
corr_spear_prot_f <- corrspear_function_ageVSabundance(dataframe = (prot_norm_meta_l %>%
                                                                  filter(Sex == "Female") %>%
                                                                  group_by(., protein)),
                                                                  analyte_type ="protein")

corr_spear_prot_pval_f <- corrspear_filtering_pval(dataframe = corr_spear_prot_f,
                                                analyte_type = "protein",
                                                adjusted_p_value_cutoff = 0.05)     

corr_spear_prot_pval_rho_f  <- corrspear_filtering_pval_rho(dataframe = corr_spear_prot_f, 
                                                         analyte_type = "protein",
                                                         adjusted_p_value_cutoff = 0.05, 
                                                         lower_spearman_rho = -0.3, 
                                                         upper_spearman_rho = 0.3)


corr_spear_pep_f <- corrspear_function_ageVSabundance(dataframe = (pep_norm_meta_l %>%
                                                                   filter(Sex == "Female") %>%
                                                                   group_by(., modifiedSequence)),
                                                                   analyte_type ="peptide")

corr_spear_pep_pval_f <- corrspear_filtering_pval(dataframe = corr_spear_pep_f,
                                                analyte_type = "peptide",
                                                adjusted_p_value_cutoff = 0.05)     

corr_spear_pep_pval_rho_f  <- corrspear_filtering_pval_rho(dataframe = corr_spear_pep_f, 
                                                         analyte_type = "peptide",
                                                         adjusted_p_value_cutoff = 0.05, 
                                                         lower_spearman_rho = -0.3, 
                                                         upper_spearman_rho = 0.3)

corr_spear_prot_pval_rho_mtx_f <- generate_prot_correlation_matrix(corr_spear_prot_pval_rho)

```


Males

```{r spearman_corr_males}

# Males
corr_spear_prot_m <- corrspear_function_ageVSabundance(dataframe = (prot_norm_meta_l %>%
                                                                  filter(Sex == "Male") %>%
                                                                  group_by(., protein)),
                                                                  analyte_type ="protein")

corr_spear_prot_pval_m <- corrspear_filtering_pval(dataframe = corr_spear_prot_m,
                                                analyte_type = "protein",
                                                adjusted_p_value_cutoff = 0.05)     

corr_spear_prot_pval_rho_m  <- corrspear_filtering_pval_rho(dataframe = corr_spear_prot_m, 
                                                         analyte_type = "protein",
                                                         adjusted_p_value_cutoff = 0.05, 
                                                         lower_spearman_rho = -0.3, 
                                                         upper_spearman_rho = 0.3)


corr_spear_pep_m <- corrspear_function_ageVSabundance(dataframe = (pep_norm_meta_l %>%
                                                                   filter(Sex == "Male") %>%
                                                                   group_by(., modifiedSequence)),
                                                                   analyte_type ="peptide")

corr_spear_pep_pval_m <- corrspear_filtering_pval(dataframe = corr_spear_pep_m,
                                                analyte_type = "peptide",
                                                adjusted_p_value_cutoff = 0.05)     

corr_spear_pep_pval_rho_m  <- corrspear_filtering_pval_rho(dataframe = corr_spear_pep_m, 
                                                         analyte_type = "peptide",
                                                         adjusted_p_value_cutoff = 0.05, 
                                                         lower_spearman_rho = -0.3, 
                                                         upper_spearman_rho = 0.3)

corr_spear_prot_pval_rho_mtx_m <- generate_prot_correlation_matrix(corr_spear_prot_pval_rho)


```

## Combine Spearman with Scatter

```{r corr_sex_subdata_points}
 # Extracting the individual normalized MS data for the above age-related markers in males and females evaluated separately
    corr_prot_prho_aging_f <- prot_norm_meta_l %>%
                                 filter(Sex == "Female") %>%
                                 filter(SampleGroup != "IEQC" & SampleGroup != "IBQC") %>%
                                 filter(protein != "sp|P00924|ENO1_YEAST") %>%
                                 subset(., protein %in% corr_targets) 

    corr_prot_prho_aging_m <- prot_norm_meta_l %>%
                                 filter(Sex == "Male") %>%
                                 filter(SampleGroup != "IEQC" & SampleGroup != "IBQC") %>%
                                 filter(protein != "sp|P00924|ENO1_YEAST") %>%
                                 subset(., protein %in% corr_targets) 

# Assigning a label in new column "Trend" to assign plot colors later.  
     corr_prot_prho_aging_f$Trend <- "Aging and Senescence"
     corr_prot_prho_aging_m$Trend <- "Aging and Senescence"

    
```


## PCA Analysis - Correlated with Age, Colored by Sex

```{r figS4a_S4b}

figS4a <- plot_pca_discrete(data_matrix_for_pca = pep_norm_matrix,
                           metadata_type_label = "Sex",
                           metadata_column_color = metadata$Sex,
                           color_palette = colpalette_sex,
                           legend_title_lower_margin = 0.1)                         

figS4a <- figS4a + ggtitle("Peptides") + labs(tag = "A") + theme(plot.title = element_text(size = 18, hjust = 0.5),
                                             legend = element_blank())


figS4b <- plot_pca_discrete(data_matrix_for_pca = prot_norm_matrix,
                           metadata_type_label = "Sex",
                           metadata_column_color = metadata$Sex,
                           color_palette = colpalette_sex,
                           legend_title_lower_margin = 0.1)      

figS4b <- figS4b + ggtitle("Proteins") + labs(tag = "B")+ theme(plot.title = element_text(size = 18, hjust = 0.5),
                                             legend = element_blank())

```

```{r figS4c}
figS4c <- plot_pca_discrete(data_matrix_for_pca = prot_corr_sig_matrix,
                           metadata_type_label = "Sex",
                           metadata_column_color = metadata$Sex,
                           color_palette = colpalette_sex,
                           legend_title_lower_margin = 0.2)      

figS4c <- figS4c + ggtitle("Proteins correlated with age, colored by sex") + labs(tag = "C")+ theme(plot.title = element_text(size = 18, hjust = 0.5))

```

## Figure S4

```{r correl_plots_sex_shared_panel}
custom_tags_fm <- c("D", "E", "F", "G", "H")#, 
                    # "I", "J", "K", "L", "M",
                    # "N", "O", "P", "Q", "R")
# custom_tags_fm <- c("D", "E", "F", "G", "H")

list_corr_prot_fm <- list()

figS4_targets <- generate_paired_sex_corr_plots(individual_data_f = (prot_norm_meta_l %>%
#figS4_lower <- generate_paired_sex_corr_plots(individual_data_f = (prot_norm_meta_l %>%
                                                                 filter(Sex == "Female")),
                                            correlation_data_f = corr_spear_prot_f,
                                            individual_data_m = (prot_norm_meta_l %>%
                                                                 filter(Sex == "Male")), 
                                           correlation_data_m = corr_spear_prot_m,
                                           targets_corr = target_proteins, 
                                           # targets_corr = corr_targets, 
                                           list_corr_prot = list_corr_prot_fm,
                                           manually_set_panel_tag = TRUE,
                                           panel_tags = custom_tags_fm)




```


```{r figS4}

figS4_targets[[1]] <- (figS4_targets[[1]] + theme(axis.title.x = element_blank()) + ylab(corr_y_axis_expr))
figS4_targets[[3]] <- (figS4_targets[[3]] + theme(axis.title.y = element_blank()) + xlab("Age (Months)"))



figS4_lower_partial <- ((figS4_targets[[1]] + figS4_targets[[2]] + figS4_targets[[3]] + figS4_targets[[4]] + figS4_targets[[5]]) + plot_layout(ncol = 5, guides = "collect"))

figS4_upper_partial <- ((figS4a + figS4b + figS4c) + plot_layout(ncol= 3, guides = "collect"))

figS4_partial <- ((figS4_upper_partial / figS4_lower_partial) +
           plot_layout(nrow = 2, guides = "collect") & theme(plot.tag = element_text(size = 25), plot.tag.position = c(0, 1)))

ggsave(file = paste(working_directory,"\\rmd_output\\plots\\",basename,filenamebase,"figS4_partial.svg", sep=""), plot=figS4_partial, width=16, height=10)


```


```{r print_figS4, fig.dim = c(16, 10)}

figS4_partial

```




<br/><br/>

# Figure S5 {.tabset .tabset-fade}

Evaluating the peptide-level Elastic Net model and what peptides are significant in that model that do not reach signficance in the protein-level model.

$~$

The Elastic Net model and OLS analyses are located in the python script by Mike Riffle (also on GitHub).


<br/><br/>


# Table 2 {.tabset .tabset-fade}

Merging results from the OLS and Spearman analyses in the python script by Mike Riffle (also on GitHub) to generate a basic version of Table 3 which was edited externally using Microsft Excel.

$~$

```{r}
extract_accession_protID <- function(dataframe){
                           # Extracting uniprot first_accession numbers
                            dataframe <- dataframe %>%
                                        mutate(first_accession = protein,
                                               first_accession = str_extract(first_accession,"(?<=\\|)[^|]+(?=\\|)"),
                                               prot_id = protein,
                                               prot_id = str_extract(prot_id,"(?<=\\|)[^|]+(?=_)"))
}

sex_spearman_filter <- sex_spearman %>%
                       #filter(fdr_diff < 0.05) %>%
                       select(protein, rho_male, rho_female, fdr_diff) %>%
                       mutate("Delta Spearman FDR < 0.05" = if_else(fdr_diff < 0.05, "Yes", "No"))
                       # mutate("Delta Spearman FDR < 0.05" = "Yes")

sex_ols_filter <- sex_ols %>%
                  #filter(q_int < 0.01) %>%
                  select(protein, beta_int, q_int) %>%
                  mutate("OLS Interaction q < 0.01" = if_else(q_int < 0.01, "Yes", "No"))
                  # mutate("OLS Interaction q < 0.01" = "Yes")

table2_partial01 <- sex_spearman_filter %>%
                    full_join(., sex_ols_filter, by = "protein")

# Simplify protein column
    table2_partial02 <- extract_accession_protID(dataframe = table2_partial01)
    
    table2_unformatted <- table2_partial02 %>%
                          filter(fdr_diff < 0.05 | q_int < 0.01) %>%
                          select(prot_id, first_accession, "Delta Spearman FDR < 0.05", rho_male, rho_female, fdr_diff, "OLS Interaction q < 0.01", beta_int, q_int) %>%
                          rename("ProteinID" = "prot_id", 
                                 "Accession" = "first_accession") %>%
                          full_join(., manual_prot_cats, by = "Accession") %>%
                          select(-ProteinID) %>%
                          mutate(rho_male = round(rho_male, 2),
                                 rho_female = round(rho_female, 2),
                                 fdr_diff = formatC(fdr_diff, format = "e", digits = 1),
                                 #fdr_diff = round(fdr_diff, 3),
                                 beta_int = round(beta_int, 3),
                                 q_int = formatC(q_int, format = "e", digits = 1))
                                 #  round(q_int, 5))

    write.csv(table2_unformatted, file = paste0(working_directory,"\\rmd_output\\tables\\",basename,filenamebase,"Table2_unformatted.csv"))
    

```


<br/><br/>

# Tables S12 & S13 {.tabset .tabset-fade}

```{r}
prot_regr_age_flag <- prot_regr_age %>%
                      mutate(feature_status = case_when(coefficient == 0 ~ "FALSE",
                                                        TRUE       ~ "TRUE"))

# Feature overlap between protein clock and significant proteins by rho and p-value in full cohort.
feature_overlap_all <- prot_regr_age_flag %>%
                         rename("protein" = "feature") %>%
                         full_join(., corr_spear_prot_pval_rho, by = "protein") %>%
                         mutate(overlap = feature_status == TRUE & p_val_adj < 0.05) %>%
                         filter(overlap == TRUE)
    

write.csv(feature_overlap_all, file = paste0(working_directory,"\\rmd_output\\tables\\",basename,filenamebase,"feature_overlap_all.csv"))
  

# Feature overlap between protein clock and significant proteins by rho and p-value in sex-difference Spearman analysis
feature_overlap_sex <- prot_regr_age_flag %>%
                       rename("protein" = "feature") %>%
                       full_join(., sex_spearman, by = "protein") %>%
                       mutate(overlap_male = feature_status == TRUE & p_male < 0.05,
                              overlap_female = feature_status == TRUE & p_female < 0.05) %>%
                       filter(overlap_male == TRUE | overlap_female == TRUE)

    write.csv(feature_overlap_sex, file = paste0(working_directory,"\\rmd_output\\tables\\",basename,filenamebase,"feature_overlap_sex.csv"))
    


```






# Package citations

Set eval to TRUE and include to TRUE to view package citations in R Markdown output.

```{r eval=FALSE, include=FALSE}

# Bioconductor packages
    citation('ReactomePA')
    citation('clusterProfiler')
    citation('org.Mm.eg.db')
    citation('UniprotR')
    citation('AnnotationDbi')
    citation('DOSE')
```

R

```{r package_citations, eval=FALSE, include=FALSE}
#Download necessary packages for analysis
    citation('car')
    citation('RColorBrewer')
    citation('dplyr')
    citation('stringr')
    citation('tidyr')
    citation('matrixStats')
    citation('ggplot2')
    citation('purrr')
    citation('ggnewscale')
    citation('ggplotify')
    citation('ggrepel')
    citation('ggpubr')
    citation('knitr')
    citation('data.table')
    citation('pheatmap')
    citation('grid')
    citation('kableExtra')
    # citation('PupillometryR')
    citation('patchwork')
    citation('cowplot')
    citation('dendextend')
    citation('viridis')
  # library('eulerr')
  citation('scales')

```


